---
title: "Lab 10"
author: "Pulindu Ratnasekera"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this lab you will write the predict method for 
mars objects. Before you start you should take a quick
look at the source code of the `predict.lm()`
function.

As in labs 8 and 9, you should be working from your `mars`
project for this lab. You will need to copy the file `testpredict.RData`
from the `Exercises/ProjectTestfiles/testthat` folder of the class GitHub 
repository to your `tests/testthat` folder.

1. Below is an implementation of the `predict.mars()` 
function. It is based on `predict.lm()`,
but with some extra fiddling to the the model matrix
right in the case of new data. (There is probably
a better way, but this is what I came up with.)
Your task is to write the function `make_B()`
that takes a design matrix `X` and a list 
of basis function specifications `Bfuncs` as input
and returns the corresponding matrix of basis functions.

```{r}
predict.mars <- function(object,newdata) {
  if(missing(newdata) || is.null(newdata)) {
    B <- as.matrix(object$B)
  }
  else {
    tt <- terms(object$formula,data=newdata)
    tt <- delete.response(tt)
    mf <- model.frame(tt,newdata)
    mt <- attr(mf, "terms")
    X <- model.matrix(mt, mf)[,-1] # remove intercept
    B <- make_B(X,object$Bfuncs)
  }
  beta <- object$coefficients
  drop(B %*% beta)
}
```

2. Create a test script `testpredict.R` in your `mars/tests/testthat` folder.
`testpredict.R` should load `testpredict.RData` and
call your `predict.mars()` function with the following
two inputs:
    1. `predict.mars(testmars)`: predictions for the
    same data used to fit the model. Compare the output
    of this call to the expected output `testpredict`.
    2. ` predict(testmars,newdata=marstestdata)`: predictions
    from the fitted model in `testmars` on "new data"
    in `marstestdata`. (This is the same data used
    to fit the model, but by passing it in as newdata 
    we test that part of the code.) Compare the output
    of this call to the expected output `testpredict`.
