---
title: "Lab 2, week 3"
author: "Pulindu Ratnasekera"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Add to shell implementation of MARS algorithm

In this lab we will investigate the possibility of having the 
`fwd_stepwise()` function and `bwd_stepwise()` functions from lab 1
use `lm()` and `step()`.

1. Simulate some test data as follows:
```{r}
set.seed(1)
x1 <- 1:100
x2 <- rnorm(100); x3 <- rnorm(100)
y <- x1+rnorm(100)
data <- data.frame(y,x1,x2,x3)
```


2. Write a couple of lines of R code to (i) fit a model with `lm()`
with response variable y and predictors x1, x2 and x3 and (ii) use
`step()` on the fitted model to do stepwise model reductions.
In your call to lm, use the `.` in your formula and pass it the 
data frame constructed in the code chunk.


3. Now try to do the same thing using your `mars()` function from lab
1 with the following modifications. 
Recall that your `mars()` function
took an R formula, `data` and `control` as inputs 
and called functions `fwd_stepwise()` and `bwd_stepwise()`.
    a. Pass the formula and data arguments to `fwd_stepwise()`. 
    Have `fwd_stepwise()` call the `lm()` function on the
    formula and data and return the resulting `lm` object.
    b. As in lab 1, your `bwd_stepwise()` function takes the output
    of `fwd_stepwise()`. Now have `bwd_stepwise()` call the 
    `step()` function on its input, and return the output of `step()`.
    c. As in lab 1, your `mars()` function returns the output of `bwd_stepwise()`. 
    d. Test your implementation with the call `mars(y~.,data=data)`.


4. Change the name of the data frame `data` in your workspace
to `testdata` are re-run mars. What happens? Use `traceback()`
to identify the function call that threw an error. What does this
tell you about where this function found the data in part 3?
