---
title: "Lab 6"
author: "Pulindu Ratnasekera"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In the recursive partitioning algorithm we had
the following constructor for S3 objects of class `region`
(see the week 7 exercises):

```{r}
new_region <- function(coords=NULL,x,y){
  if(is.null(coords)) { coords <- sapply(x,range) }
  structure(list(coords=coords,x=x,y=y),class="region")
}
```

For this lab we will define constructor, validator and 
helper functions following the conventions in the text. 
To start, we simplify the constructor to take only
the "base object" as input:

```{r}
new_region <- function(R){
  structure(R,class="region")
}
```


1. Write a validator that takes a list `R` with
elements `coords`, `x` and `y` as input and verifies that 
    (i)  `coords` is a matrix,
    (ii) `coords` has two rows and as many columns as there are columns of `x`, 
    (iii) `x` is a data frame, and 
    (iv) `y` is of type double. Use the naming convention discussed in 
    class for the validator.

2. Write a helper function that takes arguments `coords`, `x` and 
`y` (like the constructor from the week 7 exercises) and 
    (i) includes the code for handling the case where the `coords` argument is `NULL`  from the original constructor, 
    (ii) coerces `y` to a `double`,  
    (iii) calls the validator, and 
    (iv) calls the constructor to return a `region` object.
