---
title: "Lab 6, week 8"
author: "Pulindu Ratnasekera"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In the recursive partitioning algorithm we had
the following constructor for S3 region
objects. A helper function has been added,
following the convention that the helper has the
same name as the class. Also included in the code chunk is
a split method for region objects.

```{r}
new_region <- function(coords=NULL,x,y){
  if(is.null(coords)) {
    coords <- sapply(x,range)
  }
  structure(list(coords=coords,x=x,y=y),class="region")
}
region <- function(coords=NULL,x,y) {
  new_region(coords=NULL,x=x,y=y)
}
split.region <- function(R,xvar,spt){
  r1_ind <- (R$x[,xvar] <= spt)
  c1 <- c2 <- R$coords
  c1[2,xvar] <- spt; c2[1,xvar] <- spt 
  Rl <- region(c1,R$x[r1_ind,,drop=FALSE],R$y[r1_ind])
  Rr <- region(c2,R$x[!r1_ind,,drop=FALSE],R$y[!r1_ind])
  return(list(Rl=Rl,Rr=Rr))
}
#-
```


Implement region objects using R6 and S4:

1.  Create a `Region` R6 class.
    (i) Include an initializer that initializes the coords field to the 
    range of the x's if the input coords is NULL.
    (ii) Include a split method for region objects based on the 
    `split.region()` function in the above code chunk.
    (iii) Test your implementation on the x and y data shown below.
    
```{r}
set.seed(123); n <- 10
x <- data.frame(x1=rnorm(n),x2=rnorm(n))
y <- rnorm(n)
```


2. Create a `Region` S4 class with a matrix slot for coords,
a data.frame slot for x and a numeric slot for y.
    (i) Include a prototype that sets the coords to `matrix(NA,0,0)`,
    x  to NULL and y to NA_real_.
    (ii) Write a helper that creates objects of class `Region` and
    initializes the coords to the range of the x's is input is NULL.
    (iii) Define a generic split method and write a split method for `Region`
    objects. Note: your S4 generic split will over-write the S3 generic for 
    split. 
    (iv) Test your implementation on the data in the code chunk below 
    part 1.
    
