---
title: "Week 8 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. (text, ex. 14.2.1) Create a `BankAccount` R6 class that stores a balance
and allows you to withdraw and deposit money. Write an initializer 
that initializes the balance, with default balance zero. 
Test your implementation by (i) creating 
an instance of your class, and (ii) chaining a 
deposit of 100, a withdrawl of 50 and then a report of the balance.

*Solution*

```{r}
library(R6)
BankAccount <- R6Class("BankAccount", list(
  balance = NULL,
  initialize = function(balance=0){
    self$balance <- balance
    invisible(self)
  },
  deposit = function(x) {
    self$balance <- self$balance + x 
    invisible(self)
  },
  withdrawl = function(x) {
    self$balance <- self$balance - x 
    invisible(self)
  })
)
x <- BankAccount$new(balance=100)
x$deposit(100)$withdrawl(50)$balance
```

2. Create a subclass of `BankAccount` that throws an error
if you overdraft; i.e., if you attempt to withdraw more than the balance.

*Solution*

```{r}
BankAccountNoOverdraft <- R6Class("BankAccountNoOverdraft",
  inherit = BankAccount,
  public = list(
    withdrawl = function(x) {
      if(self$balance < x) stop("Error: No overdraft allowed")
      self$balance <- self$balance - x
      invisible(self)
    }
  ))
y <- BankAccountNoOverdraft$new(balance=100)
try({ y$withdrawl(200) })
```


3. Create a subclass of `BankAccount` that allows overdraft, but
charges a $10 fee.

*Solution*

```{r}
BankAccountOverdraftFee <- R6Class("BankAccountNoOverdraft",
  inherit = BankAccount,
  public = list(
    overdraftFee = 10,
    withdrawl = function(x,overdraftFee=10) {
      if(self$balance < x) {
        self$balance <- self$balance - x - overdraftFee
      } else {
        self$balance <- self$balance - x
      }
      invisible(self)
    }
  ))
y <- BankAccountOverdraftFee$new(balance=100)
y$withdrawl(200)$balance
```


4. Implement the bank account as an S4 class called `BankAccount4`.
    (i) Include a prototype that sets the balance to `NA_real_` (a missing
    value of type double).
    (ii) Write a helper that creates objects of class `BanKAccount4` and 
    allows the user to set the initial balance, with default zero.
    Have the helper function coerce the balance to a double before
    creating the object.
    (iii) Write a `show`()` method that prints the object nicely.
    (iv) Define generics and write deposit and withdrawl methods; 
    because these methods modify the balance they should 
    include calls to `validObject()`.
    (v) Use `methods()` to find all methods for objects of class 
    `BankAccount4`.

*Solution*

```{r}
# (i)
setClass("BankAccount4", 
  slots = c(
    balance = "numeric"
  ), 
  prototype = list(
    balance = NA_real_
  )
)
# (ii)
BankAccount4 <- function(balance=0){
  balance <- as.double(balance)
  new("BankAccount4",balance=balance)
}
ba <- BankAccount4(balance=100)

# (iii)
setMethod("show", "BankAccount4", function(object) {
  cat(is(object)[[1]], "\n",
      "  Balance: ", object@balance, "\n",
      sep = ""
  )
})
ba

# (iv)
setGeneric("deposit", function(x,value) standardGeneric("deposit"))
setMethod("deposit","BankAccount4",function(x,value) {
  x@balance <- x@balance + value
  validObject(x)
  x
})
setGeneric("withdrawl", function(x,value) standardGeneric("withdrawl"))
setMethod("withdrawl","BankAccount4",function(x,value) {
  x@balance <- x@balance - value
  validObject(x)
  x
})
ba <- deposit(ba,100); ba
ba <- withdrawl(ba,50); ba

# (v)
methods(class="BankAccount4")
```
