---
title: "Week 7 exercise solutions"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Using OOP terminology, what is the difference between `t.test()` and `t.data.frame()`. When is each function called?

*Solution* t.test is a generic and t.data.frame is the data.frame method
for the generic t() function. t.test is called by users. t.data.frame
is dispatched by t().

2. Describe the difference in behaviour in the following two calls to 
`mean()`. 

```{r}
set.seed(1014)
some_days <- as.Date("2017-01-31") + sample(10, 5)

mean(some_days)
#> [1] "2017-02-06"
mean(unclass(some_days))
#> [1] 17203
```

*Solution* Use s3_dispatch() to see that the first calls mean.Date(), while
the second calls mean.default. For the second, the unclassed Date
vector is numeric.

3. Refer to the lecture 7 notes. Write a validator
and helper for the node class discussed on page 14.
Have the validator check that the node's data is a
region object. The helper should take the x and y data
as input and should call the validator.

*Solution*

```{r}
# Need the constructor for region objects:
new_region <- function(coords=NULL,x,y){
  if(is.null(coords)) { coords <- sapply(x,range) }
  structure(list(coords=coords,x=x,y=y),class="region")
}
# First is the constructor from the notes. OK to redefine.
new_node <- function(data,childl=NULL,childr=NULL){
  structure(list(data=data,childl=childl,childr=childr), class="node")
}
# Validator -- lots of possibilities here
validate_node <- function(node) {
  if(!inherits(node$data,"region")) stop("node data must be a region object")
  node
}
node <- function(x,y) {
  data <- new_region(x=x,y=y)
  validate_node(new_node(data))
}
set.seed(123); n <- 10
x <- data.frame(x1=rnorm(n),x2=rnorm(n))
y <- rnorm(n)
nn <- node(x,y)
nn
```


4. Write a generic function `plot_regions` that plots regions
from the recursive partitioning algorithm in lab 2. Write methods
for this generic for objects of class tree and of class node.

*Solution* This was in the lecture slides:

```{r}
plot_regions <- function(x,...)   UseMethod("plot_regions")
plot_regions.tree <- function(tree){
  plot(tree$data$x[,1],tree$data$x[,2],xlab="X1",ylab="X2") 
  plot_regions.node(tree$childl)
  plot_regions.node(tree$childr)
}
plot_regions.node<- function(node) {
  if(is.null(node)) return(NULL)
  x <- node$data$coords[,1]
  y <- node$data$coords[,2]
  lines(c(x[1],x[2],x[2],x[1],x[1]),c(y[1],y[1],y[2],y[2],y[1]),
        col="red")
  plot_regions.node(node$childl)
  plot_regions.node(node$childr)
}
# mytree <- recpart(x,y)
# plot_regions(mytree)
```
 
5. What generics does the table class have methods for?
What generics does the ecdf class have methods for?

*Solution*

```{r}
library(sloop)
s3_methods_class("table")
s3_methods_class("ecdf")
```