---
title: "Week 7 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Using OOP terminology, what is the difference 
between `t.test()` and `t.data.frame()`. When is each function called?


2. Describe the difference in behaviour in the following two calls to 
`mean()`. 

```{r}
set.seed(1014)
some_days <- as.Date("2017-01-31") + sample(10, 5)

mean(some_days)
#> [1] "2017-02-06"
mean(unclass(some_days))
#> [1] 17203
```

3. Refer to the lecture 7 notes. Write a validator
and helper for the node class discussed on page 14.
Have the validator check that the node's data is a
region object. The helper should take the x and y data
as input and should call the validator.


4. Write a generic function `plot_regions` that plots regions
from the recursive partitioning algorithm in lab 2. Write methods
for this generic for objects of class tree and of class node.

 
5. What generics does the table class have methods for?
What generics does the ecdf class have methods for?

