---
title: "Week 6 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. The code chunk at the end of this set of exercises throws an error.
Cut-and-paste the error into a google search. Read the first few
hits and see if any are informative.

2. What function was called when the error was triggered?


3. Run traceback to see the call stack. What is the last function
that you wrote to be called? Which line of `recpart_fwd()` called 
this function?


4. Modify the code by removing the for-loop over M and setting M to 1.
Does the modified code throw an error?


5. Un-do your modification from question 4. Insert a `browser()`
at line 12 of `recpart_fwd()` (just inside the for loop over v)
and run the function.
    i. When `browser()` stops the function, print out the value of v and n. 
    ii. Step through the function with `n` a few times and then type `c` to 
    continue execution to the next call to `browser()`. Print v again. 
    iii. Use `f` to finish execution of the v-loop. Print v again. 
    Repeat a few times until the function stops. What was
    the value of v just before the crash? 
    
6. Remove your `browser()` and type `debug(recpart_fwd)`. Now run
the function under the debugger. Type out the value of a few of 
the variables as you go. When you are done, type `undebug(recpart_fwd)`
to stop debugging.

7. Insert `print()` or `cat()` statements at around line 12 of 
`recpart_fwd()` to print out the values of `v` and `n`. Re-run
`recpart_fwd()`.


```{r}
recpart_fwd <- function(y,x,Mmax){
  N <- length(y) # sample size
  n <- ncol(x) # number of predictors
  B <- init_B(N,Mmax) # Exercise: write init_B()
  splits <- data.frame(m=rep(NA,Mmax),v=rep(NA,Mmax),t=rep(NA,Mmax))
  #---------------------------------------------------
  # Looping for forward selection:
  for(M in 1:Mmax) { # contrast to indexing 2...Mmax in Friedman
    lof_best <- Inf
    for(m in 1:M) { # choose a basis function to split
      for(v in 1:n){ # select a variable to split on
        browser()
        tt <- split_points(x[,v],B[,m]) # Exercise: write split_points() 
        for(t in tt) { 
          Bnew <- data.frame(B[,(1:M)[-m]],
                        Btem1=B[,m]*(x[,v]>t),Btem2=B[,m]*(x[,v]<=t)) 
          gdat <- data.frame(y=y,Bnew)
          lof <- LOF(y~.,gdat) #  Use your LOF() from week 4
          if(lof < lof_best) { 
            lof_best <- lof
            splits[M,] <- c(m,v,t) # will
          } # end if
        } # end loop over splits
      } # end loop over variables
      n <- n-1
    } # end loop over basis functions to split
    m <- splits[M,1]; v <- splits[M,2]; t <- splits[M,3]
    B[,M+1] <- B[,m]*(x[,v]<=t)
    B[,m] <- B[,m]*(x[,v]>t)
  } # end loop over M
  return(list(B=B,splits=splits))
}
H <- function(x) { 
  return(as.numeric(x>=0)) 
}
LOF <- function(form,data) {
  ff <- lm(form,data)
  return(sum(residuals(ff)^2))
}
#-------------------------------------

#if(Mmax<2) {
#  warning("Input Mmax must be >= 2; setting to 2")
#  Mmax <- 2
# } 

init_B <- function(N,Mmax) {
  B <- data.frame(matrix(NA,nrow=N,ncol=(Mmax+1)))
  B[,1] <- 1
  names(B) <- c("B0",paste0("B",1:Mmax))
  return(B)
}
split_points <- function(xv,Bm) {
  out <- sort(unique(xv[Bm>0]))
  return(out[-length(out)])
}

set.seed(123); n <- 10
x <- data.frame(x1=rnorm(n),x2=rnorm(n))
y <- rnorm(n)
rp <- recpart_fwd(y,x,Mmax=9)
```

