---
title: "Week 2 Exercise Solutions"
author: "Brad McNeney"
output: 
  html_document:
    fig_width: 6
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Predict the outcome of the following:
```
c(1, FALSE)
c("a",1)
c(list(1),"a")
c(TRUE,1)
```

*Solution* You can see what they return by typing each line into the 
R console. The key point is that combining different data types coerces
the less flexible type to the more flexible type. For atomic
vectors, logicals are least 
flexible, then numeric, then character. As vectors, list is more 
flexible than atomic. 


2. If `vv <- list(a=1,b=2)`, why doesn't `as.vector(vv)`
work to coerce `vv` into an atomic vector?

*Solution* `vv` is already a vector

3. What do `dim()`, `nrow()` and `ncol()` return when applied to a 1-dimensional vector? What about `NROW()` or `NCOL()`?

*Solution* Try `x <- c(1,2,3); dim(x); nrow(x); ncol(x)` to see that
each returns NULL, because a 1-d vector has no dim attribute. NROW and 
NCOL treat a vector as a 1-column matrix.


4. What is `dim(cbind(A,A))` if `A = matrix(1:4,2,2)`?

*Solution* Type into the R console to find out. The cbind creates
a 2x4 matrix by putting 2 copies of A side-by-side.

5. What do the following return? Understand why. 
```
TRUE | FALSE
c(TRUE,TRUE) & c(FALSE,TRUE)
c(TRUE,TRUE) && c(FALSE,TRUE)
```

*Solution* The key here is that `&` is vectorized
while `&&` only checks the first 
element of the two vectors. 

6. What sort of object does `table()` return? What is its type? What attributes does it have? How does the dimensionality change as you tabulate more variables?

*Solution* An array with class attribute "table". The array is of type int.
The dimension increases with each variable. Tabulating 2 variables returns
a 2-d array (matrix), 3 variables returns a 3-d array, etc. Try the 
following in R

```
tt <- table(c(1,1,2,2,3,3),c(1,1,1,3,3,3),c(1,1,1,1,2,2)) 
tt
str(tt)
```

7. What happens to a factor when you modify its levels?
How do f2 and f3 differ from f1?

```
f1 <- factor(letters)
levels(f1) <- rev(levels(f1))
f2 <- rev(factor(letters))
f3 <- factor(letters, levels = rev(letters))
```

*Solution* The labels on the underlying integer values change, but
the integer values themselves don't. Try the following:

```
f1 <- factor(letters)
f1
str(f1)
levels(f1) <- rev(levels(f1))
f1
str(f1)
f2 <- rev(factor(letters))
f2
str(f2)
f3 <- factor(letters, levels = rev(letters))
f3
str(f3)
 data.frame(f1,as.numeric(f1),f2,as.numeric(f2),f3,as.numeric(f3))
```

You can see that f2 and f1 print the same but the integer-to-label
correspondence is backwards. Conversely, f3 and f3 have the same 
integer-to-label correspondence, but the labels are in reverse order.

8. Fix each of the following common data frame subsetting errors:
```
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]
```

*Solution* 

```
mtcars[mtcars$cyl == 4,]
mtcars[-(1:4),]
mtcars[mtcars$cyl <= 5,]
mtcars[mtcars$cyl ==4 | mtcars$cyl == 6 | ]
```

9. Consider the following data frame and tibble. Suppose
you want to extract the number from the first row and first column and 
multiply this number by the vector `1:3`. How would
you do this for the data frame? How would you do this
for the tibble?
```{r}
dd <- data.frame(x=1:3,y=c("a","b","c"))
library(tibble)
tt <- as.tibble(dd) # Note: we are supposed to use as_tibble() now
```

*Solution*

```{r}
dd[1,1]*(1:3)
as.numeric(tt[1,1])*(1:3)
```


10. Read the help file `help("%in%")`. Copy the following data frame into R.
```{r}
students <- data.frame(
           major=c("Stat","Stat","Econ","Bus","Phys","ActSci","ActSci"),
           year=c(3,4,4,5,3,3,4))
```
Use `%in%` to create a logical vector of students who major in Stat
or ActSci and use this to subset `students` to Stat or ActSci majors.

*Solution*

```{r}
students[students$major %in% c("Stat","ActSci"),]
```