---
title: "Week 5 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## I. Recursive partitioning by forward selection

* In this exercise you will write a version of 
recursive partitioning by forward selection
following Algorithm 1 of Friedman (1991), with 
a few notational changes that will make 
the transition to the MARS forward selection 
algorithm easier.

* The following code gives the framework for the function
`recpart_fwd()`.
You are asked to write several of the support 
functions described below the code chunk.

* You need to understand `recpart_fwd()` so that you 
can modify it to the MARS forward stepwise algorithm. 



```{r}
recpart_fwd <- function(y,x,Mmax){
  #---------------------------------------------------
  # Error checking:
  #---------------------------------------------------
  # Initialize:
  N <- length(y) # sample size
  n <- ncol(x) # number of predictors
  B <- init_B(N,Mmax) # Exercise: write init_B()
  splits <- data.frame(m=rep(NA,Mmax),v=rep(NA,Mmax),t=rep(NA,Mmax))
  #---------------------------------------------------
  # Looping for forward selection:
  for(M in 1:Mmax) { # contrast to indexing 2...Mmax in Friedman
    lof_best <- Inf
    for(m in 1:M) { # choose a basis function to split
      for(v in 1:n){ # select a variable to split on
        tt <- split_points(x[,v],B[,m]) # Exercise: write split_points() 
        for(t in tt) { 
          Bnew <- data.frame(B[,(1:M)[-m]],
                        Btem1=B[,m]*(x[,v]>t),Btem2=B[,m]*(x[,v]<=t)) 
          gdat <- data.frame(y=y,Bnew)
          lof <- LOF(y~.,gdat) #  Use your LOF() from week 4
          if(lof < lof_best) { 
            lof_best <- lof
            splits[M,] <- c(m,v,t) 
          } # end if
        } # end loop over splits
      } # end loop over variables
    } # end loop over basis functions to split
    m <- splits[M,1]; v <- splits[M,2]; t <- splits[M,3]
    B[,M+1] <- B[,m]*(x[,v]<=t)
    B[,m] <- B[,m]*(x[,v]>t)
  } # end loop over M
  return(list(B=B,splits=splits))
}
LOF <- function(form,data) {
  ff <- lm(form,data)
  return(sum(residuals(ff)^2))
}
#-------------------------------------
```

Exercises:

1. Write a code snippet that checks 
whether the input `Mmax` is greater than or equal to 2. If not, 
throw a warning and set Mmax to 2.
2. Write the function `init_B()`, which takes the sample 
size `N` and max number of basis functions `Mmax` as input
and returns a data frame with 1's in the first column, NAs in
all others, and column names B0,B1,...,BMmax.
3. Write `split_points()`. This one is different from the 
one you wrote in the week 4 exercises. Here we take predictor
variable $x_v$, and basis function $B_m$ as input. The eligible split 
points are $x_v$ values such that basis function $B_m(x)>0$.
Return ordered unique values, and remember to exclude the 
largest possible value.
4. Test `recpart_fwd(y,x,Mmax=2)`, with your functions from 
the exercises, runs on the following test data. 
These are the same as in the week 4 exercises.
See if you can find these first two splits on the drawing from week 4: 

[image](exrecpart.png)

```{r}
set.seed(123); n <- 10
x <- data.frame(x1=rnorm(n),x2=rnorm(n))
y <- rnorm(n)
# rp <- recpart_fwd(y,x,Mmax=2)
```


* You will create tests of your implementation in lab 3.

## II. MARS forward selection

* Over reading break, think about how to 
implement the MARS forward selection algorithm (Algorithm 2). 
As a modification of recursive partitioning, recall the 
following differences:

1. Replace the mirror-image indicator basis functions with 
mirror-image hinge functions.
2. Do not remove parent basis function $B_m(x)$ after it
is split. We are therefore adding two basis functions in 
each outer-loop iteration.
3. Only split basis function $B_m(x)$ into $B_m(x)h(t-x_v)$
and $B_m(x)h(x_v-t)$ for variables $x_v$ not already involved in $B_m$.
