---
title: "Week 9 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**IMPORTANT** The following exercises should be done from 
the `mars` RStudio project that you created in lab 5. (If you haven't started the R package yet, please
do so as soon as possible.) The root folder of this project should be
called `mars`. If you see error messages like
```
Error: Could not find package root. Is your working directory inside a package?
> 
```
you are not working in your `mars` project.
After last week's lab 7 your `mars` project should have working versions of the helper, validator
and constructor for `mars.control` objects, and a working version of 
`fwd_stepwise()`.

1. Use `usethis::use_vignette("mars")` to create a  skeleton vignette
for your package.

2. Use `usethis::use_testthat()` to create a skeleton test suite 
for your package. 

3. Use `usethis::use_data_raw("marstestdata")` to create a `data-raw` folder
and skeleton R script for creating the `marstestdata` dataset. 

4. Copy-and-paste the following into the `marstestdata.R` script 
created in question 3. The dataset is of 
size N=100 with n=10 explanatory variables, and a response 
variable that depends on only the first two explanatory variables.
Source the script to create the `marstestdata` dataset for your
package. Note: Leave
`usethis::use_data(marstestdata,overwrite=TRUE)` as the last
line of the script -- this saves the dataset in the `data` folder.

```{r}
# Set a seed for reproducibility
set.seed(123)
# Parameters
N <- 100; n <- 10
knot1 <- -0.5; knot2 <- 0; knot3 <- 0.5
beta1 <- 3; beta2 <- 5
error.SD <- 0.1
# Simulate x's
x <- matrix(rnorm(N*n),ncol=n)
# Simulate response variable
h <- function(x,s,t) { 
  return(pmax(0,s*(x-t)))
}
lin.pred <- beta1*h(x[,1],+1,knot1) + beta2*h(x[,2],-1,knot2)*h(x[,1],+1,knot3)
y <- lin.pred + rnorm(n,sd=error.SD)
# Package response and explanatory variables as a data frame
marstestdata <- data.frame(cbind(y,x))
names(marstestdata) <- c("y",paste0("x",1:n))
```

5. To document the `marstestdata` dataset, copy the file
`Exercises/ProjectTestfiles/data.R` from the class GitHub repository
to your `mars/R` folder. Look at the `data.R` file and make
sure you understand its contents.

6. Copy the files `testmc.RData` and `testfwd_stepwise.RData` 
from the `Exercises/ProjectTestfiles` folder of the class
GitHub repository to your `mars/tests/testthat`
folder. Create test scripts `testmars_control.R`, and
`testfwd_stepwise.R` in your `mars/tests/testthat` folder. 
    * `testmars_control.R` should load `testmc.RData`, 
    call your `mars.control()` function with input `Mmax=10`, 
    and compare the output to the output `testmc` in `testmc.RData`
    * `testfwd_stepwise.R` should load `testfwd_stepwise.RData`,
    call your `fwd_stepwise()` function with the inputs 
    `testy`, `testx`, and `testmc` from `testfwd_selection.RData`,
    and compare the output to `testfwd` from `testfwd_selection.RData`.

7. Use `devtools::test()` to run the tests. Both tests should pass.

8. (Optional, but recommended) 
Commit the files that were created or modified in these exercises to your
project folder on GitHub.