---
title: "Week 3 exercises"
author: "Brad McNeney"
date: '2019-01-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Control flow

1. What type of vector does each of the following return?

```{r}
ifelse(TRUE, 1, "no")
ifelse(FALSE, 1, "no")
ifelse(NA, 1, "no")
```

2. Re-write the following using `switch`

```{r}
IQR_mid <- function(x) mean(quantile(x,c(.25,.75)))
cc <- function(x,method) {
  if(method=="mean") {
    mean(x)
  } else if(method=="median") {
    median(x)
  } else if(method=="IQR_mid") {
    IQR_mid(x)
  } else stop("centring method ",method," not implemented")
}
set.seed(123)
x <- c(-3,rnorm(100),1000)
cc(x,"mean")
cc(x,"median")
cc(x,"IQR_mid")
try(cc(x,"cat"))
```

3. Rewrite the following function so that 
it uses a `while()` loop instead of the 
`for()` loop and `break` statement. Your while-approach
will not require the `maxit` upper limit on the 
number of iterations.
```{r}
rtruncNormal <- function(thresh = 2, maxit=1000) {
  x<-NULL
  for(i in 1:maxit) {
    xnew <- rnorm(n=1)
    if(xnew>thresh) {
      break
    }
    x <- c(x,xnew)
  }
  x
}
set.seed(1234)
rtruncNormal()
```


## Functions

4. The following code chunk is typed into the R Console.
    - What is the output of the function call `f(5)`?
    - What is the enclosing environment of `f()`?
    - What is the enclosing environment of `g()`?
    - What search order does R use to find the 
    value of `x` when it is needed in `g()`?
```
x <- 1
f <- function(y) {
  g <- function(z) {
    (x+z)^2
  }
  g(y)
}
f(5)
```


