---
title: "Week 11 exercises"
author: "Brad McNeney"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Install the R package `nyclfights13`, and use the following 
commands to save a subset of the `flights` dataframe as a .csv file.

```{r}
library(nycflights13)
library(tidyverse)
set.seed(123)
flights <- sample_n(flights,size=5000) %>% select(dep_delay,carrier,dep_time)
write.csv(flights,file="flights.csv")
```


Write a Python script that reads in the data, drops observations
with missing values and fits a regression
of `dep_delay` on `carrier` and `dep_time`. Save 
the residuals and fitted values from the regression.
In R, plot the residuals *vs* fitted values.
Compare this to a plot of residuals *vs* fitted values when
you fit the model in R.

Notes:

* The variable `carrier`
is a categorical variable that will need to be converted to 
dummy variables before you fit the model.
* Depending on your approach, reticulate may 
return the fitted values and residuals as matrices.

Some suggested sources of information are:

* The reticulate website at https://rstudio.github.io/reticulate/.
The script in the first code chunk reads data in from a
file called flights.csv,
extracts a subset of rows, a subset of variables and
then drops observations with missing values.
* The pandas documentation for the `get_dummies()` function
https://pandas.pydata.org/docs/reference/api/pandas.get_dummies.html
* The `lec11_2.py` Python script, which uses the `linear_model()`
function from `sklearn` to fit a regression.

Don't forget to load the conda environment (or similar) that you
installed pandas and sklearn into; e.g.,

```{r}
library(reticulate)
use_python("/home/mcneney/miniconda3/bin/python") # modify as necessary
use_condaenv("r-reticulate") # modify as necessary
```
