---
title: "Stat 360 Week 1 exercises"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. On which line does a get copied in the following example?
```{r}
a <- c(1, 5, 3, 2)
b <- a
b[[1]] <- 10
```

*Solution* Line 3:

```{r}
a <- c(1, 5, 3, 2)
tracemem(a)
b <- a
tracemem(b)
b[[1]] <- 10
```

2. In the following code chunk, does `x` get copied? Does `ll` get copied? Does  `ee` get copied?
```{r}
x <- rnorm(100); y <- rnorm(100)
ll <- list(x=x,y=y)
ll$x <- 1:100
ee <- rlang::env(x=x,y=y)
ee$x <- 1:100
```

*Solutions* Use `tracemem()` on `x`, `y` and `ll` to see that only `ll` 
gets copied. (Recall that you can't put a trace on the environment `ee`.)

3. Find the size of the objects x and y in the following code chunk.
Which is smaller? If instead of vectors from 1 to 10 they were vectors
from 1 to 1 million, which would be smaller?

```{r}
x <- c(1,2,3,4,5,6,7,8,9,10)
y <- 1:10
```

*Solution* Use `lobstr::obj_size()` to see that `x` is actually 
smaller! However, for vectors of length 1 million `y` would be smaller. 
You can see this if, for example, you generate 1 million random numbers:

```{r}
lobstr::obj_size(x)
lobstr::obj_size(y)
x <- rbinom(1e6,10,.5)
y <- 1:1e6
lobstr::obj_size(x)
lobstr::obj_size(y)
```