---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 5'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---


## R packages, CRAN and GitHub

*  CRAN and the R package system is the key to R's success, allowing contributions
from hundreds of scientists outside the "R core team".
* CRAN has strict quality-control checks and requirements.
* It is possible to distribute R packages *via* GitHub without the 
CRAN checks.

## R packages

* What is an R package?
    * An easy-to-install collection of R functions, documentation and 
    example data.

* Why make an R package?
    * Share your work with others
    * Get credit for your work
    * Make it easier for you to use your own work.

## Making R packages

* The definitive source for making R packages is the 
"Writing R Extensions" document, which you can find at
https://mirror.rcg.sfu.ca/mirror/CRAN/doc/manuals/r-release/R-exts.html

* A minimal R package is a folder containing DESCRIPTION and NAMESPACE files
and an R subfolder.
* Other possible subfolders are data, demo, exec, inst, man, po, src, tests, tools and vignettes. 
* Other possible files in the directory are INDEX, configure, cleanup, LICENSE, LICENCE and NEWS. 

## Getting Started

* A helper function called `package.skeleton()` in the `utils` 
package that comes with R takes a package name and list of 
R objects as arguments and sets up directories, files
and skeleton help files.

\small

```{r}
rm(list=ls()) # clear workspace
source("../mars.R") # location of my mars source file
# package.skeleton(name="mars",ls()) # Can't be re-run
```


##

\footnotesize

```
Creating directories ...
Creating DESCRIPTION ...
Creating NAMESPACE ...
Creating Read-and-delete-me ...
Saving functions and data ...
Making help files ...
Done.
Further steps are described in './mars/Read-and-delete-me'.
```


## 

\footnotesize

```
$ more mars/Read-and-delete-me 
* Edit the help file skeletons in 'man', possibly combining help files
  for multiple functions.
* Edit the exports in 'NAMESPACE', and add necessary imports.
* Put any C/C++/Fortran code in 'src'.
* If you have compiled code, add a useDynLib() directive to
  'NAMESPACE'.
* Run R CMD build to build the package tarball.
* Run R CMD check to check the package tarball.

Read "Writing R Extensions" for more information.

```

## devtools

* A package called `devtools` provides more help.
    * `devtools` was started by Hadley Wickham and has since expanded
    into a collection of tools for developing R packages, described
    in a work-in-progress book called "R Packages" 
    https://r-pkgs.org/index.html
    * We'll skim Chapter 2, which provides an overview of functionality.
* In addition to automating more of the process, devtools 
uses `roxygen2` to create documentation from 
structured comments in your R source files.
    * C++ programmers will recognize the approach from Doxygen

## Getting started with `create_package()`

* Call `create_package()` to initialize an R package directory and 
new R project
* Recommended that this new directory **not** be part of an
existing R project or be under version control.
* Launches a new RStudio session in the newly-created directory/project
    * Switch to this new session.
    * If needed, call `use_git()` to initialize a git repository

##

\scriptsize

```{r}
# install.packages("devtools")
library(devtools)
create_package("/Users/mcneney/MyRpackages/mars")
```

## Start your R source files

* Copy your R source files to the R sub-folder.
    * For now there is just  mars.R, but later we will have predict.mars.R, plot.mars.R, etc.
* Call `load_all()` to load the source into your R session.
    * Rather than just source in your R code, `load_all()`
    loads the R functions as a package.
    * Gives a better sense of how the code will behave when loaded 
    by a user.

```{r}
library(devtools) # call in new R session
# load_all()
```


## check

* R CMD check from the command line
* `check()`
* Will throw a warning of a non-standard licence 

## Edit DESCRIPTION file

* Add your name, collaborators, etc. 

## Add a licence

* Helper functions add a copy of the relevant license to your 
package main directory and update the Licence field of the 
DESCRIPTION file.


\small

```{r}
# use_gpl3_license()
```

## Add documentation comments to your source file.

* Open your .R file(s), place your cursor
in one of the functions and click `Code->Insert Roxygen Skeleton` 
to insert a skeleton of the roxygen2-style comments above 
the function. 

\footnotesize

```
#' Title
#'
#' @param formula 
#' @param data 
#' @param control 
#' @param ... 
#'
#' @return
#' @export
#'
#' @examples
```

##

\footnotesize

```
#' Multivariate Adaptive Regression Splines (MARS)
#'
#' @param formula an R formula
#' @param data a data frame containing the data for the model
#' @param control an object of class 'mars.control'
#' @param ... other arguments -- currently not used
#'
#' @return an object of class 'mars'
#' @export
#'
#' @examples
#' mm <- mars(wage ~ age,data=ISLR::Wage)
#' @import stats
#' @import ISLR
```

## Call `document()`

* Calling `document()` will create the .Rd file from your comments, and
add `mars` as an "export" and `stats` and `ISLR` as "imports" in the mars NAMESPACE file.
    * See below for Imports

\small

```
> document()
Updating mars documentation
Loading mars
Writing NAMESPACE
Writing mars.Rd
> load_all()
Loading mars
> ?mars
Rendering development documentation for 'mars'
```

## Add other packages you depend on

* Reference: R packages, Chapter 8
* Your package's reliance on other packages can be as 
    * "Depends" -- you expect that users will always want 
    to call library(package) when loading your package -- not common these days
    * "Imports" -- you need to call functions from the NAMESPACE of another
    package. Add these to the DESCRIPTION file with `use_package()`.
    * "Suggests" -- Packages that are not crucial, but helpful (e.g.,
    used in examples). Add these to the DESCRIPTION file yourself
    * "Enhances" -- Packages that are enhanced by yours -- not common
    
```{r}
# use_package("stats")
# use_package("ISLR") # used in mars example
# document()
```


## check

* Call `check()` to check your package using the same checks
as `R CMD check` from the command line.
* Should return no errors, warnings or notes.

\small

```{r}
# check()
```

## install

* Call `install()` to install the package in your R library.

\small

```{r}
# install()
```

## Topics not covered

* You have enough to get you started, but you will need to learn more
as you go ...
* Our MARS implementation will use the S3 object oriented programming 
system. 
    * You will need to read about documenting classes, generics and methods in Chapter 10, SS 7
* Your final project document will include a "vignette"
    * Chapter 11 
* You will need to write "unit tests" using the `testthat` package
    * Chapter 12

