---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 3'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

# Control Flow

## Control Flow

- Reading: text, chapter 5
- if/if-else, ifelse, switch
- for
- while
- break

## `if` and `if`-`else` 


- `if` tests a condition and executes code if the condition is true. Optionally, can couple with an `else` to specify code to execute
when condition is false.

\scriptsize

```{r}
if("cat" == "dog") {
  print("cat is dog")
} else {
  print("cat is not dog")
}
```

## `if` returns a value

- The body of the if-else can evaluate expressions and store results,
but note that if-else also returns a value.

\scriptsize

```{r}
cnd <- if("cat" == "dog")  "cat is dog" else "cat is not dog"
cnd
```

## `if` expects a single logical

- most other inputs will cause an error
- logical vectors will not throw an error, but `if` will only
use the first element

\scriptsize

```{r}
try(if("cat") print("cat"))
if(c("cat"=="dog","cat" == "cat")) print("hello world") 
```

## `ifelse()`: vectorized `if`

- `ifelse()` can handle logical vectors
- syntax is condition, what to return if expression true, what
to return if expression false

\scriptsize
```{r}
x <- 1:10
ifelse(x %% 2 == 0, "even","odd")
```

## `switch`

- If you have multiple conditions to check, consider `switch` instead
of repeated if-else; e.g. \scriptsize
    * `if(x==1) "cat" else if(x==2) "dog" else if (x==3) "mouse"`

\scriptsize

```{r}
x <- 2 # numeric argument
switch(x,"cat","dog","mouse") # evaluate the x'th element
x <- "dog" #character argument
switch(x,cat="hi cat",dog="hi dog",mouse="hi mouse",
       warning("unknown animal")) # if we make it to the last condition
switch("kangaroo",cat="hi cat",dog="hi dog",mouse="hi mouse",
       warning("unknown animal"))
```


## `for` loops

- Example:

\scriptsize

```{r}
n <- 10; nreps <- 100; x <- vector(mode="numeric",length=nreps)
for(i in 1:nreps) {
  # Code you want to repeat nreps times
  x[i] <- mean(rnorm(n))
}
summary(x)
print(i)
```

## `for` loop index set

\small

- Index sets such as `1:n` are most common, but 
can be almost any atomic vector.

```{r}
ind <- c("cat","dog","mouse")
for(i in ind) {
  print(paste("There is a",i,"in my house"))
}
```

## `seq_along` 

* A common use of for loops is to iterate over elements
of a vector, say x. 
* Creating the index set `1:length(x)` will not be what you 
expect when x has length 0 (e.g., x is NULL).
* Instead use `seq_along()`

\scriptsize

```{r}
x <- NULL
for(i in 1:length(x)) print(x[i])

for(i in seq_along(x)) print(x[i])
```



## `while` loops

\small

- Use a `while` loop when you want to continue until
some logical condition is met.

```{r}
set.seed(1)
# Number of coin tosses until first success (geometric distn)
p <- 0.1; counter <- 0; success <- FALSE
while(!success)  {
  success <- as.logical(rbinom(n=1,size=1,prob=p))
  counter <- counter + 1
}
counter
```

## `break`

- `break` can be used to break out of a `for` or `while` loop.

```{r}
for(i in 1:100) {
  if(i>3) break
  print(i)
}
```

## `repeat`

* `repeat` continues indefinitely until it encounters a `break`
* The text considers `repeat` to be the most flexible of `for`, `while`
and `repeat`.


# R Functions

## R function fundamentals

* Reading: text sections 6.1 and 6.2
\small
- In R, functions are objects with three essential components: 
    - the code inside the function, or `body`,
    - the list of arguments to the function, or `formals`, and
    - an `environment` that contains all objects defined
    in the function.
    
- Functions can have other attributes, but the above three are 
essential.

## Example function 
\scriptsize

```{r}
f <- function(x) {
  return(x^2)
}
f
```

## The function body
\small

- This is the code we want to execute. 
- When the end of a function is reached without a call to 
`return()`, the value of the last line is returned.
    - So in our example function, we could replace `return(x^2)` 
    with just `x^2.
- Use `body()` to see the body of a function.

\scriptsize

```{r}
body(f)
```

## The function formals


- These are the arguments to the function. 
- Function arguments can have default values and/or be defined
in terms of other arguments.

\scriptsize

```{r}
f <- function(x=0) { x^2} 
f <- function(x=0,y=3*x) { x^2 + y^2 }
f()
f(x=1)
f(y=1)
```

##

* Use `formals()` to see the formals of a function and 
their default values.

\scriptsize

```{r}
formals(f)
```


## Argument matching when calling a function
\footnotesize

- When you call a function, the arguments are matched first by 
name, then by "prefix" matching and finally by position:
```{r}
f <- function(firstarg,secondarg) {
  firstarg^2 + secondarg
}
f(firstarg=1,secondarg=2)
f(s=2,f=1)
f(2,f=1)
f(1,2)
```

## The function environment

- R creates an environment (with `rlang::env()`) within each function call to hold its variables.
- Initially includes the formals, but variables created within the function are also stored in this
environment

\scriptsize

```{r}
f <- function(x) {
  y <- x^2
  ee <- environment() # Returns ID of environment w/in f
  print(ls(ee)) # list objects in ee
  ee
}
# After a function call its environment is usually discarded
my_ee <- f(1:3) # but not if you bind it to a name
my_ee$new_var <- 100
x <- 1:100
ls(my_ee)
```


##

\scriptsize

```{r}
environment(f)
```

## Enclosing environments
\small

- Our function `f` was defined in the global environment,
`.GlobalEnv`,
which "encloses" the environment within `f`.
- If `f` needs a variable and can't find it within
`f`'s environment, it will look for it in the enclosing 
environment, and then the enclosing environment of `.GlobalEnv`,
and so on. 
- The `search()` function lists the heirarchy of 
environments that enclose `.GlobalEnv`. 
```{r}
search()
```
- To facilitate this search, each environment includes 
a pointer to its enclosing environment.

## R packages and the search list

\small

- Use the `library()` command to load packages.
- When we load a package
it is inserted in position 2 of the search list, 
just after `.GlobalEnv`.

\footnotesize

```{r}
# install.packages("hapassoc")
library(hapassoc) 
search()
```

## Detaching packages

- Detach a package from the search list with `detach()`

\footnotesize

```{r}
detach("package:hapassoc")
search()
```

## Package namespaces
\small

- Package authors create a list of objects
that will be visible to users when the package
is loaded. This list is called  the package namespace.
- You can access functions in a package's namespace 
without loading the package using
the `::` operator.
```{r}
set.seed(321)
n<-30; x<-(1:n)/n; y<-rnorm(n,mean=x); ff<-lm(y~x)
car::sigmaHat(ff)
```
- Doing so does not add 
the package to the search list.


## Up next

* Reading: Text, sections 6.4-6.8