---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 4'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

## Digging deeper into functions

* Reading: Text, sections 6.4-6.8
* Topics:
    * more on scoping (finding objects)
    * lazy evaluation and variable arguments with ...
    * exiting a function
    * prefix, infix, replacement and special function forms


# Scoping

## Lexical scoping in R

* We have already touched on the essence of scoping in R: When
a computation needs an object we start by looking in 
the current environment, and then search successive 
enclosing environments.
* More formally R has four rules:
    * Name masking
    * Functions versus variables
    * A fresh start
    * Dynamic lookup
    
## Name masking

* A consequence of the search order for objects is that
names defined *inside* a function mask names defined *outside*.
    * This is 

\small

```{r}
x <- y <- 200
z <- 30 # defined in global environment
f <- function() { # f's env enclosed by global
  x <- 100 # defined in f's environment
  y <- 20
  g <- function() { #g's env enclosed by f's
    x <- 10 # defined in g's environment
    c(x,y,z)
  }
  g()
}
f()
```

## Each function call gets a new environment

* All objects created within the function
disappear when the function exits.

\small

```{r}
x <- 100
f <- function(){
  print(environment())
  x <- x+1
  x
}
f()
f()
```

## Dynamic lookup

* Be aware that functions only look for objects when run (dynamic lookup), 
not when created (static lookup). 
* If a function gets an object from an enclosing environment,
it will return different results whenever the object in 
the enclosing environment changes.
    * This may be what you intend, but it's also a common source
    of errors. What if in the following I meant to define y
    in f() but forgot?

\small

```{r}
y <- 100
f <- function(x) {
  x + y
}
f(1)
y <- 200
f(1)
```


# Lazy evaluation and ...

## Lazy evaluation

* Function arguments are only evaluated when needed.
    * The text describes how lazy evaluation is implemented
    (Section 6.5.1), but we will not discuss the details.

\small

```{r}
f<-function(xx,yy) {
  xx
}
f(1) # no value for yy, but OK since yy not used
try(f(yy=1)) # xx is needed
```


## Variable arguments with ...

* The special function argument `...` (dot-dot-dot) allows
a function to take any number of arguments.
* A typical use is to pass these to another function, as in 
the following example.

\small

```{r,fig.height=4}
myplot <- function(x,...) {
  plot(x,x^2,...) # pass any args not named x to plot
}
myplot((-5:5),col="red",pch=16)
```

# Exiting a function

## Exiting a function

* Functions can exit explicitly with `return()`
or implicitly, where the last expression in the
function is its return value
* When a function returns, explicitly or implicitly, 
the default is to print the return value. 
    * You can suppress this with `invisible()`.

\small

```{r}
ff <- function(x) { x }
ff(1)
ff_invis <- function(x) { invisible(x) }
ff_invis(1) # but x <- ff_invis(1) same as x <- ff(1)
```

## Signalling conditions

* Functions can signal error, warning or message
conditions with `stop()`, `warning()` and `message()`,
respectively.
    * `stop()` stops execution, `warning()`
    and `message()` don't
* These signals can be "handled" by ignoring them
    * ignore errors with `try()`
    * ignore warnings with `suppressWarnings()`
    * ignore messages with `suppressMessages()`
* or implementing a custom handler that over-rides
the default behaviour of a condition
    * see Chapter 8 of the text for more on 
    handling conditions 
* We restrict attention to (i) signalling and (ii) 
cleaning up any changes to the R session before exiting.



## `stop()`

* If your function encounters an error, use `stop()` 
to stop and print an error message, also called
"throwing" an error.

\small

```{r}
centre <- function(x,method) {
  switch(method,mean=mean(x),median=median(x),
         stop("method ",method," not implemented"))
}
try(centre(1:10,"mymean"))
```


## `warning()` 

* If you suspect an error but can proceed without
stopping, throw a `warning()` instead.

\small

```{r}
centre <- function(x,method) {
  switch(method,mean=mean(x),median=median(x),
         {warning("\nmethod ",method,
                  " not implemented, using mean\n"); 
           mean(x)})
}
centre(1:10,"mymean")
```

## `message()`

* If you don't think the condition warrants a warning,
you can issue a message.

```{r}
centre <- function(x,method) {
  switch(method,mean=mean(x),median=median(x),
         {message("\nmethod ",method,
                  " not implemented, using mean\n"); 
           mean(x)})
}
centre(1:10,"mymean")
```


## Cleaning up with exit handlers

* An R session has a "global state" of options and
parameters that control default behaviour.
    * type `options()` or `par()` to see some of these
* If your function temporarily modifies the global state, 
you can use an exit handler to re-set, even if your
function stops.
    * Use `add=TRUE` to add more than one handler.
    

\small

```{r}
rplot <- function(y,x){
  opar <- par(mfrow=c(2,2))
  on.exit(par(opar),add=TRUE) # add=TRUE not nec. in this ex.
  plot(lm(y~x)) #could throw an error
}
y <- rnorm(100); x <- rnorm(10) # different length
try(rplot(y,x)) # Fails, but re-sets par mfrow
```

# Function forms

## Function forms

* We have been writing "prefix" functions, with a function
name followed by arguments.
* Other forms are "infix", "replacement" and "special".
* We will cover each form very briefly; see the text,
section 6.8 for more details.

## Infix functions

* An infix function has two arguments and
is called by putting the name
between arguments, as in `x+y`.
    * `x+y` calls `+` as ``` `+`(x,y)```
    * `+` and `-` are special infix functions that
    can be called with only one argument
    * You can define your own infix function by 
    enclosing the function name in `%`.

\small

```{r}
`%-%` <- function(set1,set2){
  setdiff(set1,set2)
}
s1 <- 1:10; s2 <- 4:6
s1 %-% s2 # same as `%-%`(s1,s2)
```

## Replacement functions

* Replacement functions are called to change values.
    * For example, change values of attributes of objects
* Must have arguments `x` and `value`, and must return the modified object.
* They are made to look like prefix functions, and 
may have prefix counterparts.

\footnotesize

```{r}
x <- c(a=1,b=2)
names(x)
names(x) <- c("aa","bb")
x
x <- `names<-`(x,c("aaa","bbb"))
x
```

##

* You can write your own relacement functions if 
you end the function name with `<-` 

\small

```{r}
`st360names<-` <- function(x,value){
  names(x) <- paste0(value,"360",names(x))
  x
}
st360names(x) <- c("a","b")
x
```

## Special functions

* Examples: subset `[` and extract `[[`, control flow
`if`, `for`,etc. 
* Key point: These are functions,
and it is sometimes useful to know their names so
that we can get help or
use them like any other prefix function.

\small

```{r}
dd <- data.frame(x=1:2,y=3:4)
`[[`(dd,1) # compare to dd[[1]]
dd <- `[[<-`(dd,1,value=5:6) #cf dd[[1]] <- 5:6
dd
```

##

* It can be useful to know functions by name so that
we can call them in `lapply`-like functions.

\small

```{r}
sapply(dd,`[[<-`,2,value=10)
```