---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 8'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

## More object-oriented programming in R

* Last time:
    * base objects vs OO objects
    * OOP with S3 in R
* Today:
    * Brief introductions to OOP with R6 and S4
    * Reading: skim chapters 14 and 15 of Advanced R by Wickham
    

# R6

## Introduction to R6

* R6 is an "encapculated" OOP system, so methods are bundled
with objects, rather than being functions designed to act on objects.
    * R6 methods are called by `object$method()`, rather
    than `generic.objectclass()`
* R6 objects are implemented using environments and so can 
be modified in place.
* R6 will look familiar to programmers coming from another language. 

\small

```{r}
#install.packages("R6")
library(R6)
```


## Defining classes and methods with R6

* Use `R6Class()` to create a class and its methods.

* Two important arguments to `R6Class()` are the classname
argument and the public argument.
    * classname is self-explanatory
    * public specifies the methods and fields that are the
    public interface of the object. Methods access components
    of the object with `self`.


## Example from the text

\small

```{r}
Accumulator <- R6Class(classname="Accumulator", 
  public = list(
    sum = 0,
    add = function(x = 1) {
      self$sum <- self$sum + x 
      invisible(self)
      }
    )
)
# Accumulator
x <- Accumulator$new() # create an Accumulator object
x$add(1) # method 
x$add(2)$add(3)$add(4) # methods can be "chained"
x$sum # field
```

## Side-effect methods

* `$add()` is called for the "side-effect" of modifying the 
sum 
* Side-effect methods should return invisibly. Otherwise
the object is printed when the method is called.

\small

```{r}
Accumulator <- R6Class(classname="Accumulator", 
  public = list(
    sum = 0,
    add = function(x = 1) {
      self$sum <- self$sum + x 
      self
      }
    )
)
# Accumulator
x <- Accumulator$new() 
x$add(1) 
```


## ... reset ...

\small

```{r}
Accumulator <- R6Class(classname="Accumulator", 
  public = list(
    sum = 0,
    add = function(x = 1) {
      self$sum <- self$sum + x 
      invisible(self)
      }
    )
)
```

## initialize and print methods

* Will make your class easier to use.
    * initialize is a constructor that over-rides 
    the default new method and allows users to initialize
    an instance of the class with data values.
    
\small

```{r}
# version without initialize and new
Person <- R6Class("Person", list(
  name = NULL,
  age = NA
  ))
brad <- Person$new()
brad$name = "Brad"
brad$age = 54
brad
```

##

\footnotesize

```{r}
# version with initialize and print
Person <- R6Class("Person", list(
  name = NULL,
  age = NA,
  initialize = function(name, age = NA) {
    stopifnot(is.character(name), length(name) == 1)
    stopifnot(is.numeric(age), length(age) == 1)
    self$name <- name
    self$age <- age
  },
  print = function(...) {
    cat("Person: \n")
    cat("Name:", self$name, "\n")
    cat("Age:", self$age, "\n")
    invisible(self)
  }
))
brad <- Person$new("Brad", age = 54)
brad
```

## Inheritance

* Use `inherit` to create a child class that inherits methods
and fields from a  parent (super) class
* You can add or over-ride methods/fields in the child 

\small

```{r}
AccumulatorChatty <- R6Class("AccumulatorChatty", 
  inherit = Accumulator,
  public = list(
    add = function(x = 1) {
      cat("Adding ", x, "\n", sep = "")
      super$add(x = x) # use the superclass implementation of add
    }
  )
)
x2 <- AccumulatorChatty$new()
x2$add(10)$add(1)$sum
```

## `class()` and `names()`

* You can use `class()` and `names()` to query an R6 object.

\small

```{r}
class(x2)
names(x2) 
```

## Making copies 

* R6 objects are implemented as environments. 
    * Objects are modified in place.
    * The usual way of making copies in R with `<-` 
    does not work:

\small

```{r}
x3 <- x2 # Are we copying x2?
x3$add(100)
x3$sum
x2$sum # !!
```

## clone

* Make copies with the `$clone()` method.

\small

```{r}
x3 <- x2$clone()
x3$add(-100)
x3$sum
x2$sum
```

## R6 topics not covered

* Private and active fields (Section 14.3)

* More on unexpected behaviour of R6 classes (Section 14.4)

* R6 *versus* the built-in reference classes (RC) system
(Section 14.5)


# S4

## Introduction to S4

* S4 is a formal functional OOP system with strict rules for
creating classes, generics and methods.
* Also has a more advanced implementation of inheritance/dispatch.
* Down-side is that it has a steeper learning curve than S3.

* Terminology: S4 objects have "slots", accessed by `@`.
    * Similar in function to list elements in most S3 classes,
    which are accessed by `$`.
    
* S4 is implemented in the `methods` package, which is loaded
automatically in every R session. 
    * However, the text recommends explicitly loading methods

## Creating classes

* Use `setClass` to create a class and the `new()` method
to create objects of that class.

\small

```{r}
library(methods)
setClass("Person", 
  slots = c(
    name = "character", 
    age = "numeric"
  )
)
brad <- new("Person", name = "Brad", age = 54)
```

## Class prototype

* In addition to the class and slot names, you should provide a 
prototype for your class. 
    * The prototype specifies default values for the slots

\tiny

```{r}
setClass("Person", 
  slots = c(
    name = "character", 
    age = "numeric"
  ), 
  prototype = list(
    name = NA_character_,
    age = NA_real_
  )
)
brad <- new("Person", name = "Brad")
str(brad)
```

##

* You can use `is()` to see an S4 object's class, and @ or `slot()` to
access slots.
    * `@` is equivalent to `$` and `slot()` is equivalent to `[[`,
    and for most purposes they are equivalent to each other.

\tiny

```{r}
is(brad)
brad@name
slot(brad,"name")
brad@name <- "Brad McNeney"
brad
```


## Inheritance

* The `contains` argument specifies a parent class to inherit slots 
and methods from. 

\tiny

```{r}
setClass("Employee", 
  contains = "Person", 
  slots = c(
    boss = "Person"
  ),
  prototype = list(
    boss = new("Person")
  )
)
brad <- new("Employee",name="Brad",boss=new("Person",name="Catherine"))
is(brad,"Employee")
is(brad,"Person")
```



## Helpers

* Just as with S3, you should write a user-friendly helper
to create objects of your class.
* The helper can perform checks, coerce data to correct types, etc.
* Give the helper the same name as the class

\small

```{r}
Person <- function(name, age = NA) {
  age <- as.double(age)
  new("Person", name = name, age = age)
}
Person("Brad")
```

## Validators

* For more complicated checks, write a validator with
`setValidity()`
* `setValidity()` takes a class and a function that returns TRUE if 
the input is valid, and a character vector describing the problem if not:

\tiny

```{r}
setValidity("Person", function(object) {
  if (length(object@name) != length(object@age)) {
    "@name and @age must be same length"
  } else {
    TRUE
  }
})
```

##

\small

```{r}
# new("Person",name="Brad",age=54:55)
# new("Employee",name=c("Brad","McNeney"))
```

##  Generics and methods

* Example: Write accessor functions for users to get and set data in 
your class.
    * Users shouldn't use `@`, and you shouldn't on other developers' classes
    * Write generics with `setGeneric()` and methods with `setMethod()`.
    * Setting values has the potential to create invalid objects. 
    Can call `validObject()` to ensure a valid object.

\tiny


```{r}
# Note: Don't use {} in the function definition of setGeneric.
# get values with a prefix function
setGeneric("age", function(x) standardGeneric("age"))
setMethod("age", "Person", function(x) x@age)
# set values with a replacement function
setGeneric("age<-", function(x, value) standardGeneric("age<-"))
setMethod("age<-", "Person", function(x, value) {
  x@age <- value
  validObject(x) # check object validity
  x
})
age(brad) <- 55
age(brad)
try({ age(brad) <- 55:56 })
```

## Signature

* The signature argument of `setGeneric()` specifies which arguments
are used for method dispatch.
    * Default is all arguments. 
* The second argument of `setMethod()` is also 
called signature, and specifies
the classes that the method applies to.
* S4 allows for generics and methods to dispatch on multiple
classes.
    * Can get quite confusing.
    * See Section 15.5 of text if interested.
  
## 

\small

```{r}
setGeneric("age<-",function(x,value,...,verbose=TRUE) standardGeneric("age<-"),
           signature = "x") # dispatch on first arg only
setMethod("age<-", "Person", function(x,value,...,verbose=TRUE) {
  x@age <- value
  if(verbose) cat("Setting age to",value,"\n")
  x
})
age(brad) <- 56
```

## show method

* The `show()` method is the S4 equivalent of print.
    * It should have one argument.

\small

```{r}
setMethod("show", "Employee", function(object) {
  cat(is(object)[[1]], "\n",
      "  Name: ", object@name, "\n",
      "  Age:  ", object@age, "\n",
      "  Boss: ", object@boss@name, "\n",
      sep = ""
  )
})
brad
```


## List methods

* Use `methods("generic")` or `methods(class = "class")` to see all 
methods for a given generic or class.

\small

```{r}
methods("age")
methods(class="Employee")
methods(class="Person")
```

## Topics skimmed or not covered.

* Method dispatch, section 15.5
* Interfacing S4 and S3, section 15.6.