---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 3'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

# Control Flow

## Control Flow

- Reading: text, chapter 5
- if, while, break, FIX

## `if` and `if`-`else` 

\small

- `if` tests a condition and executes code if the condition is true. Optionaly, can couple with an `else` to specify code to execute
when condition is false.

```{r}
if("cat" == "dog") {
  print("cat is dog")
} else {
  print("cat is not dog")
}
```


## `for` loops

\small

Example:

```{r}
n <- 10; nreps <- 100; x <- vector(mode="numeric",length=nreps)
for(i in 1:nreps) {
  # Code you want to repeat nreps times
  x[i] <- mean(rnorm(n))
}
summary(x)
print(i)
```

## `for` loop index set

\small

- Index sets of the form `1:n` are most common, but 
can be almost any atomic vector.

```{r}
ind <- c("cat","dog","mouse")
for(i in ind) {
  print(paste("There is a",i,"in my house"))
}
```

## `while` loops

\small

- Use a `while` loop when you want to continue until
some logical condition is met.

```{r}
set.seed(1)
# Number of coin tosses until first success (geometric distn)
p <- 0.1; counter <- 0; success <- FALSE
while(!success)  {
  success <- as.logical(rbinom(n=1,size=1,prob=p))
  counter <- counter + 1
}
counter
```

## `break`

- `break` can be used to break out of a `for` or `while` loop.

```{r}
for(i in 1:100) {
  if(i>3) break
  print(i)
}
```

# R Functions

## R functions

* Reading: text, chapter 6
\small
- In R, functions are objects with three essential components: 
    - the code inside the function, or `body`,
    - the list of arguments to the function, or `formals`, and
    - a data structure called an `environment` which is like a 
    map to the memory locations of all objects defined
    in the function.
    
- Functions can have other attributes, but the above three are 
essential.

## Example function 
\scriptsize

```{r}
f <- function(x) {
  return(x^2)
}
f
```

## The function body
\small

- This is the code we want to execute. 
- When the end of a function is reached without a call to 
`return()`, the value of the last line is returned.
    - So in our example function, we could replace `return(x^2)` 
    with just `x^2.

## The function formals
\small

- These are the arguments to the function. 
- Function arguments can have default values, as in:
```{r}
f <- function(x=0) { x^2} 
```
- Argument defaults can be defined in terms of other arguments:
```{r}
f <- function(x=0,y=3*x) { x^2 + y^2 }
f()
f(x=1)
f(y=1)
```

## Argument matching when calling a function
\footnotesize

- When you call a function, the arguments are matched first by 
name, then by "prefix" matching and finally by position:
```{r}
f <- function(firstarg,secondarg) {
  firstarg^2 + secondarg
}
f(firstarg=1,secondarg=2)
f(s=2,f=1)
f(2,f=1)
f(1,2)
```

## The function environment
\small

- The environment within a function is like a map to the 
memory locations of all its variables.
- The function arguments are "passed by value", meaning that
a copy is made and stored in the function's environment.
- Variables created within the function are also store in its
environment
```{r}
f <- function(x) {
  y <- x^2
  ee <- environment() # Returns ID of environment w/in f
  print(ls(ee)) # list objects in ee
  ee
}
f(1) # function call
```

## Enclosing environments
\small

- Our function `f` was defined in the global environment,
`.GlobalEnv`,
which "encloses" the environment within `f`.
- If `f` needs a variable and can't find it whithin
`f`'s environment, it will look for it in the enclosing 
environment, and then the enclosing environment of `.GlobalEnv`,
and so on. 
- The `search()` function lists the heirarchy of 
environments that enclose `.GlobalEnv`. 
```{r}
search()
```
- To facilitate this search, each environment includes 
a pointer to its enclosing environment.

# R packages

## Loading packages
\small

- Use the `library()` command to load packages.
- When we load a package
it is inserted in position 2 of the search list, 
just after `.GlobalEnv`.

\footnotesize

```{r}
# install.packages("hapassoc")
library(hapassoc) 
search()
```

## Detaching packages

- Detach a package from the search list with `detach()`

\footnotesize

```{r}
detach("package:hapassoc")
search()
```

## Package namespaces
\small

- Package authors create a list of objects
that will be visible to users when the package
is loaded. This list is called  the package namespace.
- You can access functions in a package's namespace 
without loading the package using
the `::` operator.
```{r}
set.seed(321)
n<-30; x<-(1:n)/n; y<-rnorm(n,mean=x); ff<-lm(y~x)
car::sigmaHat(ff)
```
- Doing so does not add 
the package to the search list.