---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 2'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

## Reading

* Text, chapters 3 and 4


# R Data Structures

## R Data Structures

- Fundamentally, all common data structures in R are vectors, which can 
be "atomic" or "list".
- R has no true scalars; e.g., in `x<-1`, `x` is a vector of length one.
- Use `str()` to see the structure of an object

## Types of objects

- All R objects have a "type", that describes how it is stored
in computer memory.
- Common types we will encounter are "logical", "integer",
"double", "character" and "list".
    * Find the type of an object with `typeof()`.
    
    
\scriptsize

```{r}
x <- 6 # stores as double by default
typeof(x)
y <- 6L # The "L" suffix forces storage as integer  
typeof(y)
```


## Type *versus* Mode

- In addition to the type of an object, there is its
"mode". 
- The mode of an object is generally the same as its type,
but the modes are coarser.
    - For example, integer and double types are both of mode
    "numeric".
- I don't understand the need for mode.
    - The only reason I mention it is that the `str()` function 
    sometimes reports the mode of an object, rather than
    its type, so we will frequently see reports of numeric objects.

\scriptsize

```{r}
mode(x)
mode(y)
```


# Vectors 

## Vectors
\small

- Vectors can be either atomic or list
    - The elements of an atomic vector must be the same type.
    - Lists can be comprised of multiple data types
- Empty vectors can be created by the `vector()` function:

\scriptsize

```{r}
# help("vector")
avec <- vector(mode="numeric",length=4)
lvec <- vector(mode="list",length=4)
```

\normalsize

- Data vectors can be created with `c()` or `list()`:

\scriptsize

```{r}
avec <- c(54,210,77)
lvec <- list(54,210,77,c("grey","thin"))
```


## Combining vectors

- Use `c()` to combine vectors

\scriptsize

```{r}
c(avec,c(100,101))
c(lvec,TRUE)
```

## Vector attributes

- Vectors have a type and length and, optionally, attributes such as names. 
    * As we have seen, we find the type of an object with `typeof()`.
    * Find the length of a vector with `length()`.

\footnotesize

```{r}
typeof(avec)
length(avec)
str(avec)
```

##

\scriptsize

```{r}
typeof(lvec)
length(lvec)
names(lvec) = c("age","weight","height","hair")
str(lvec)
```

\normalsize

- We can specify element names when creating a vector; e.g.:

\scriptsize

```{r}
lvec <- list(age=54,weight=210,height=77,hair=c("grey","thin"))
```

## Factors

- The statistical concept of a factor is important in experimental 
design. 
- Factors are implemented in R as atomic vectors with attributes
`class` and `levels`:

\scriptsize

```{r}
trt <- factor(c("drug1","placebo","placebo","drug2"))
attributes(trt)
str(trt)
```

##

- The levels are coded numerically (1, 2 and 3)
with assigned labels ordered alphabetically
("drug1", "drug2" and "placebo") by default.
- You can specify an order to the factors with 
the `level` argument:

\scriptsize

```{r}
trt <- factor(c("drug1","placebo","placebo","drug2"),
              levels=c("placebo","drug1","drug2"))
trt
```


## More on object class

- You can create your own `class` for an object. 
- Such "meta-data" can be used to tell R how to handle the 
object; e.g., how to print it, summarize it, etc.

\scriptsize

```{r}
class(lvec) <- "prof"
lvec
```


## Subsetting vectors and extracting elements

- Subset with `[` or by name:

\scriptsize

```{r}
lvec[c(1,3)] # same as lvec[c("age","height")]
```

\normalsize

- Extract individual elements with `[[`, or `$` for named objects:

\scriptsize

```{r}
lvec[[4]]
lvec$hair
```

## Subsetting factors

- Subsetting may remove all instances of a level, but the 
level will be retained in the data structure

\scriptsize

```{r}
trt[1:3]
```

\normalsize

- If subsetting is intended to remove a level of the factor, use `drop=TRUE`

\scriptsize

```{r}
trt[1:3,drop=TRUE]
```

## Subsetting and assignment

- You can combine subsetting and assignment to change 
the value of vectors. 

\scriptsize

```{r}
avec
avec[2] <- 210
avec
```

## Assignment to vector elements

* To assign to a vector element, it is clearer to use `[[` rather than `[`.
    - Also, for lists, assignment with `[` requires that the replacement 
    element be of length 1; `[[` 
    does not have this restriction
    
\scriptsize

```{r}
lvec[3:4] <- c("Hi","there")
lvec[3:4]
```

##

\scriptsize

```{r}
lvec[4] <- c("All","of","this")
lvec[4] # Only used first element of replacement vector
lvec[[4]] <- c("All","of","this")
lvec[3:4]
```


## Coercion: atomic vectors to lists

- Atomic vectors can be coerced to lists with `as.list()`:

\scriptsize

```{r}
avec = c(age=54,weight=210,height=77)
avec
as.list(avec)
```

## Coercion: lists to atomic vectors

- Lists can be "flattened" into atomic vectors with `unlist()`:

\scriptsize

```{r}
unlist(lvec)
```

\small

- Notice how the numeric values are coerced to the more flexible 
character type. 
- The order of flexibility, from least to most, is logical,
integer, numeric, character.

## Coercion: factors to atomic vectors

- We saw how to use `factor()` to coerce an atomic vector to a factor.
- Use `as.vector()` to coerce a factor back to an atomic vector. 
- The result is a character vector. You may need to use `as.numeric()` to coerce to numeric, if required.

\scriptsize

```{r}
a <-factor(c(2,1,1,2))
as.vector(a)
as.numeric(as.vector(a))
```


# Matrices and data frames

## Matrices and data frames

- Matrices are implemented as atomic vectors with a "dim" attribute of
length 2  (number of rows, number of 
columns).
- As an atomic vector, elements of a matrix must all be of the same type.
- Data frames are lists where each list element
has the same length. Thus data frames can include columns
of varying type.

## Matrices

- Matrices can be created with the `matrix()` function as in 
\scriptsize

```{r}
A <- matrix(1:4,nrow=2,ncol=2) 
A
```

\normalsize

- Here `1:4` is the same as `c(1,2,3,4)`

##

- The default is to read the data vector into the matrix column-by-column. To 
read row-by-row instead use the `byrow=TRUE` argument:

\scriptsize

```{r}
A <- matrix(1:4,nrow=2,ncol=2,byrow=TRUE) 
A
```

## Combining matrices


- Combine matrices with `rbind()` and `cbind()`:

\scriptsize

```{r}
rbind(A,matrix(c(5,6),nrow=1,ncol=2))
cbind(A,A)
```

## Matrix attributes

- Matrices have a type, dimension and optional attributes such as dimnames (row and column names).

\scriptsize

```{r}
typeof(A)
dim(A)
```

##

\scriptsize

```{r}
colnames(A) <- c("var1","var2")
rownames(A) <- c("subj1","subj2")
A
str(A)
```

## Subsetting matrices 
\footnotesize

- Subset with `[` and a comma to separate rows from columns:
```{r}
A[1,1]
A[1,]
A[,1]
```
- When a subsetting operation leads to a vector, the dimension 
of the object is "dropped" from 2 to 1. To prevent this use
`drop=FALSE`:
```{r}
A[1,,drop=FALSE]
```


## Extracting elements from matrices 
\footnotesize

- Can use `[[` to extract elements, but this is not necessary 
because of the way subsetting to a single element drops to 
a vector of length 1 by default:
```{r}
A[[1,1]]
A[1,1]
```

## Coercion: Matrices to/from vectors
- We have already seen how `matrix()` coerces a vector to a matrix
- `as.vector()` applied to a matrix removes the dim attribute and 
creates a vector by concatenating columns:
```{r}
as.vector(A)
```

##  Data frames
\small

- Data frames (class `data.frame`) are the usual way to 
store data in R.
    - Rows are intended to be observational units, columns variables
    - Implemented as a list (columns are list elements), but also behave like a matrix in terms of combining and subsetting.
- Create with `data.frame`:

\footnotesize

```{r}
set.seed(1)
n <- 4
x <- 1:n; y <- rnorm(n,mean=x,sd=1) # multiple commands separated by ;
dd <- data.frame(x=x,y=y) # like making a list
str(dd)
```

## Subsetting and combining data frames like a list

\scriptsize

```{r}
dd$x 
dd[[1]]
```

## Subsetting and combining data frames like a matrix

\scriptsize

```{r}
dd[1:2,]
zz = data.frame(z=runif(4))
cbind(dd,zz)
```


# Logical  and relational operators
## Logical operators
\small

- The basic logical operators are described in `help("Logic")`.
- `!` is NOT
- `&` and `&&` are AND, with `&` acting vector-wise and `&&` acting 
on scalars
- `|` and `||` are OR, with `|` acting vector-wise and `||` acting 
on scalars
- Make sure you understand the following:
```{r}
x <- c(TRUE,TRUE,FALSE); y <- c(FALSE,TRUE,TRUE)
!x ; x&y ; x&&y ; x|y ; x||y
```

## 

- Notice how `&&` and `||` act on the first element of the vectors
`x` and `y` and ignore all the rest.


## Relational operators
\small

- Relational operators can be used to compare values in atomic vectors
    - See `help("Comparison")`
- `>` is greater than, `>=` is greater than or equal
- `<` is less than, `<=` is less than or equal
- `==` is equal and `!=` is not equal
- Make sure you understand the following:
```{r}
x <- 1:3; y <- 3:1
x>y ; x>=y ; x<y ; x<=y ; x==y ; x!=y
```


## Subsetting vectors with logical expressions
\footnotesize

- Can subset with logicals and `[`:
```{r}
avec
avec>100
avec[avec>100]
avec[avec>54 & avec<100]
```

## Subsetting matrices with logical expressions

- Can also subset matrices, but results may not be as expected:
```{r}
A
A>1
A[A>1] # coerces to a vector
```

## Subset and assign with logical expressions
\footnotesize

- Combine subset and assign to change the value of objects
```{r}
A[A>1] <- 9
A
```
- In the above substitution, the vector 9 is shorter than the 
three elements in `A>1` so R "recycles" the 9 three times. 

## Be careful about recycling:

\footnotesize

```{r}
A[A>1] <- c(-10,10) # Throws a warning 
A # R used c(-10,10), then just the -10 
```

# Aside: Special values

## Missing values
\small

- R has a special data code for missing data: `NA`
- Test for and set missing values with `is.na()`
```{r}
avec
is.na(avec)
is.na(avec) <- 2 
avec
```



## Infinite and undefined values

\footnotesize

- R has a special codes for infinite 
values (`Inf`) and undefined values (`NaN`).

- Test for `Inf` and `NaN` with `is.infinite()` and `is.nan()`.
```{r}
ii <- 1/0 ; nn <- 0/0
ii
is.infinite(ii)
nn
is.nan(nn)
```

## The null object

- The null object, `NULL`, is an un-typed no-value object.
    - Test for `NULL` with `is.null()`
    - `NULL` can be used to initialize objects 
    that will be created through combining, rbinding, etc.
    
\footnotesize

```{r}
x <- NULL; is.null(x)
x <- c(x,1); x <- c(x,2); x 
# etc., or as a loop (more on these later)
x <- NULL
for(i in 1:2) {
  x <- c(x,i)
}
x
```

 
