---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 11'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

##  R and Python: References 

* The reticulate website: https://rstudio.github.io/reticulate/
* Python setup: https://docs.python.org/3/using/index.html
* Python tutorial: https://docs.python.org/3/tutorial/

## Calling Python from R 

* Use cases:
    * Tap into a growing set of tools for data science, such as
    scikit-learn https://scikit-learn.org/stable/ , keras https://keras.io/ ,
    ...
    * Workflow requires substantial computations in **both** languages;
    e.g., fit a neural network in Python, plot results using ggplot2

* Not recommended:
    * Using R/RStudio as a development environment for Python.
    * Better to go all-in with Python and use Jupyter Notebooks,
    Spyder, etc. as your IDE.

## Prerequisites

* Install and load the reticulate package

\small

```{r}
# install.packages("reticulate")
library(reticulate)
```

\normalsize

* Install Python
    * See the Python setup link on the references slide,
    or the Python distributions on the next slide
    * Warning for Mac users: Macs with the Apple M1
    processor can run both M1 and x86 (Intel processor)
    binaries. Make sure your R and Python installations 
    are for the same processor, or you will get errors 
    when you try to start Python from R.
* Install the Python packages that you need.
    * More on this below.
    
## Python packages

* Python packages are like R packages.
* Install with `pip` or `conda`.
    * So far I've only been able to find a miniconda installer
to work with my M1 version of R, so I'll discuss `conda`-based installs.
* conda comes with the Python distributions anaconda
and miniconda https://docs.conda.io/projects/conda/en/latest/user-guide/install
    * miniconda is smaller and faster to install, but you
    will need to install packages yourself
    * anaconda includes all the python packages you will 
    need for data science but takes more disk space and
    probably includes stuff you will never use
* Once you have `conda` install packages 
from the command line with `conda install <package>`.
    

## Python environments

* Python environments, like RStudio projects, are used to compartmentalize
your work with Python. 
    * A complete Python installation, including its own Python 
    executable and packages.
    * Create from the command line with `conda create --name <env_name>`
    * Then "activate" with `conda activate <env_name>` and 
    "de-activate" with `conda deactivate`.

## Installing packages with reticulate, Part I

* See
https://rstudio.github.io/reticulate/articles/python_packages.html

* We'll first install into the "base" Python environment

\footnotesize

```{r}
library(reticulate)
#conda_install(packages=c("pandas","scikit-learn")) # commented out to avoid re-doing
```

## Installing packages with reticulate, Part II

* Now install into another Python environment
* Note: Not currently working on my system, but it used to!

\footnotesize

```{r}
library(reticulate)
#conda_create("r-reticulate") # commented out to avoid re-doing every time I knit
#conda_install("r-reticulate", c("pandas","scikit-learn")) # commented out to avoid re-doing
```

## Using a conda environment

* For each R session in which you want to use your conda environment:

\small 

```{r}
library(reticulate)
use_condaenv("r-reticulate",required=TRUE)
# Can then call py_config() to see which Python version is being used
```


## Python embedded in RMarkdown

* Example from https://pandas.pydata.org/pandas-docs/stable/user_guide/10min.html

\small

```{python}
# code chunk header is ```{python} rather than ```{r}
import numpy as np
import pandas as pd
df = pd.DataFrame(np.random.randn(3,4),columns=['A','B','C','D'])
df
```


## Importing Python packages (modules)

* You can also import Python packages into R and call their functions
directly.

\small

```{r}
npr <- import("numpy.random") 
pd <- import("pandas") # import is from reticulate
df <- pd$DataFrame(npr$randn(3L,4L),columns=c('A','B','C','D'))
df
```

## Notes

* Access Python functions from an imported package with `$`.
* The `randn()` function requires integer arguments -- have to 
use `3L` and `4L` to pass integers.
    * reticulate converts R vectors of length 1 to Python scalars.
    * In general reticulate will try to convert to/from appropriate
    data types. See the list at https://rstudio.github.io/reticulate/index.html#type-conversions
* I used the numpy random number generator, but passed column names
as an R character vector.
    * reticulate converts this to a python list.


## Sourcing Python scripts

* Source with `source_python()`. By default,
objects created by the Python script are 
made available as R objects in your workspace,
and also in the hidden environment `py`.
* Pass an environment if you want to store somewhere
else.

\scriptsize

```{r}
# source_python("lec11_1.py") # default
my_py_env <- new.env()
source_python("lec11_1.py",envir=my_py_env)
ls(my_py_env)
my_py_env$df
```

## Another example

\tiny

```{r,warning=FALSE,message=FALSE,fig.height=4}
source_python("lec11_2.py")
MSE
ddat <- data.frame(Y=diabetes_y,diabetes_X)
library(ggplot2)
ggplot(ddat,aes(x=X1,y=Y)) + geom_point() + geom_smooth()
```

## Python REPL

* You can also start the Python interpreter and compute
interactively.
    * Useful for debugging your Python scripts

\small

```{r}
# repl_python()
# type your Python commands
# objects will be available in R through py object
# exit to quit
```

