---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 10'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

##  References: 

* Chapter 25 of Advanced R by Wickham.
* The Rcpp website: http://www.rcpp.org/
* Rcpp gallery: https://gallery.rcpp.org/
* Rcpp quick reference https://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf


## Calling C++ from R with Rcpp

* R is written in C, and so in principle it is possible to write C/C++ code
that calls R and R code that calls C/C++.
    * Using R "internals" directly is complicated.
    * A base R function for passing R objects to C/C++ is `.Call`.
    * See the chapter "System and foreign language interfaces"
    in the "Writing R Extensions" manual. 
* Rcpp provides a more user-friendly interface with C++.
    * Allows you to write C++ functions that can be called from R
    * Use it to speed up code that runs too slowly in R.
    * Of the 17345 packages on CRAN, 2385 (or more) use Rcpp
* This is a brief intro; much more info at http://www.rcpp.org/ and 
elsewhere on the internet.

## Use cases

* Unavoidable loops (can't avoid by vectorizing).

* Many (e.g., millions) of function calls, such as in a recursive algorithm.
    * Less overhead in C++ compared to R.

* Require advanced data structures available in a C++ library,
such as the Standard Template
Library( STL), but not in R.

* And many more ...

## Prerequisites

* Install and load the Rcpp package

\small

```{r}
library(Rcpp)
```

\normalsize

* Install a working C++ compiler. To get it:
    * On Windows, install Rtools.
    * On Mac, install Xcode.
    * On Linux, `sudo apt-get install r-base-dev`
    

## First example

* Example  from section 25.2 of text:

\small

```{r}
cppFunction('int add(int x, int y, int z) {
  int sum = x + y + z;
  return sum;
}')
# add works like a regular R function
add
add(1, 2, 3)
```

* Rcpp (i) compiles the C++ code (note lag) and (ii) constructs the 
R function `add()` that will call this compiled code.
* The above defines a function "inline"; it is also possible to "source" 
C++ code (more later).

## The plan


* Start simple and work up, writing functions with:
    * no inputs and a scalar output
    * scalar input and scalar output
    * vector input and scalar output
    * vector input and vector output
    * matrix input and vector output
* Keep an eye out for differences, such as the need to declare the 
type of objects.

## No inputs, scalar output

\small

* R:

```{r}
one <- function() 1L
```

* Rcpp:

```{r}
cppFunction('int one() {
  return 1;
}')
one()
```

## C++ differences

* Don’t use assignment to create functions.

* Declare the type of output the function returns; e.g., 
a scalar integer (`int`).


* C++ has true scalars, with types `double`, `int`, `String`, and `bool`.

* A return statement is **required**

* Every statement is terminated by a `;`.


## Scalar input and output

* A scalar version of the sign() function which returns 1, 0 or $-1$ for positive, zero or negative input:

\small

```{r}

signR <- function(x) {
  if (x > 0) { 1 } else if (x == 0) { 0 } else { -1 }
}

cppFunction('int signC(int x) {
  if (x > 0) {
    return 1;
  } else if (x == 0) {
    return 0;
  } else {
    return -1;
  }
}')
signC(-100)
```

## Notes

* In addition to declaring the type of the output, we must
declare the type of the input.
* Logicals and `if`-`else` are the same.

## Vector input and scalar output


\footnotesize

```{r}
sumR <- function(x) {
  total <- 0
  for (i in seq_along(x)) {
    total <- total + x[i]
  }
  total
}
cppFunction('double sumC(NumericVector x) {
  int n = x.size();
  double total = 0;
  for(int i = 0; i < n; i++) {
    total += x[i];
  }
  return total;
}')
```


##


\small

```{r}
set.seed(1)
x <- rnorm(1e5)
bench::mark(sumR(x),sumC(x),sum(x))
```


## Notes

* The input (R vector) in this example is of type `NumericVector`.
    * Other R vector types are `IntegerVector`, `CharacterVector`, 
    and `LogicalVector`. 
* The `.size()` method of a vector returns the length as an integer.
* Notice the syntax of `for()`: `for(initial condn; check condn; increment)`. 
    * In this case we initialise by creating variable i with value 0. 
    * Before each iteration we check that `i < n`, and terminate if not. 
    * After each iteration, increment i by one, using `++`.

* C++ uses zero-based indexing, 0,...n-1 (!!!)

* `+=` increments "in-place"


## Vector input, vector output

\small

```{r}
pdistR <- function(x, ys) { sqrt((x - ys) ^ 2) }
cppFunction('NumericVector pdistC(double x, NumericVector ys) {
  int n = ys.size();
  NumericVector out(n);

  for(int i = 0; i < n; ++i) {
    out[i] = sqrt(pow(ys[i] - x, 2.0)); //  pow() vs ^{}
  }
  return out;
}')
pdistC(10,6:15)
```


* `NumericVector out(n)` is a constructor.
* Copy an existing vector with the `clone()` function; 
e.g., `NumericVector zs = clone(ys)`.

## Using sourceCpp

* For functions of more than a few lines it is more convenient to 
define the function(s) in a source file and use `sourceCpp()`
to link it/them to R. 
* Source files must end in `.cpp` and include

```
#include <Rcpp.h>
using namespace Rcpp;
```

* If you use the `File -> New File -> C++ file` feature of RStudio 
it will add the above lines for you.

```{r}
sourceCpp("lec10_1.cpp") # See source file
timesTwo(42)
```


## List input, including S3 classes

* Rcpp provides wrappers for lists/data frames, functions, and attributes.
* Generally more useful for output than input, because C++ needs to
know classes of list components in advance.
* For use as input, you can convert components to C++ equivalents with `as()`.
    * See the source file for the following example.

\footnotesize

```{r}
sourceCpp("lec10_2.cpp") # see source file
mod <- lm(mpg ~ wt, data = mtcars)
mpe(mod)
```


## Functions

* So far our R <-> C++ interface has been R -> C++, but we can 
also call R functions from C++.
* Use type `Function` to input R functions and type `RObject` to
hold general input/output.

https://gallery.rcpp.org/articles/r-function-from-c++/

\footnotesize

```{r}
sourceCpp("lec10_3.cpp") # see source file
set.seed(123)
x <- rnorm(100)
callFunction(x,fivenum)

callWithOne(function(x) x+1)

fit <- lm_in_C(formula(mpg~disp),mtcars,lm)
summary(fit)$coef
```


## A warning from the expeRts


*  From https://gallery.rcpp.org/articles/r-function-from-c++/
\begin{quote}
Calling a function is simple and tempting. It is also slow as there are overheads involved. And calling it repeatedly from inside your C++ code, possibly buried within several loops, is outright silly. This has to be slower than equivalent C++ code, and even slower than just the R code (because of the marshalling of data). Do it when it makes sense, and not simply because it is available.
\end{quote}

## Attributes

* Get and set R object attributes, such as name and class,
with the `.attr()` method for R vector types from Rcpp.
   *  Rcpp also provides a .names() method specifically for the names attribute. 
   
\footnotesize

```{r}
sourceCpp("lec10_4.cpp") # see source file
attribs()
```

## 
## Further Reading

* Missing values, section 25.4

* Rcpp gives us access to the data structures and algorithms 
provided by C++ libraries like the Standard Template Library.
    * See Section 25.5 for examples.
    * See also "The Algorithm Design Manual" (https://www.algorist.com/),
    or the online text and 
    course "Algorithms" (https://algs4.cs.princeton.edu/home/)
* Two case studies are given in Section 25.6
* Using Rcpp and C++ code in an R package is discussed in Section 25.7


