---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: "Multivariate Adaptive Regression Splines (MARS)"
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
    fig_width: 4
    fig_height: 2.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```


## Example Data

\footnotesize

```{r}
library(tidyverse)
library(ISLR)
data(Wage) # help(Wage) for info
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=.2) + geom_smooth()
```

## 

\footnotesize

```{r}
library(earth)
ee <- earth(wage ~ age, data=Wage)
Wage <- mutate(Wage,pwage = predict(ee))
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=.2) + geom_smooth()+
  geom_line(aes(y=pwage),color="red")
```

## 

\footnotesize

```{r}
summary(ee)
```


##

\footnotesize

```{r}
ggplot(Wage,aes(x=age,y=wage)) + geom_point(alpha=.2) +
  geom_line(aes(y=pwage),color="red") +
  geom_vline(xintercept=38) +
  geom_vline(xintercept=63)
```


## Hinge functions

* The points 38 and 63 are "knots" where the piece-wise 
linear function changes slope.
* The piece-wise linear fit is a linear model in a constant 
term (intercept) 
and two "hinge" functions, $h(38-age)$ and $h(age-63)$, where
$$
h(x) = \max(0,x)
$$
* Hinge functions $h(x-c)$ and $h(c-x)$ are called mirror image.
    * Exercise: Plot two mirror-image hinge functions for 
    `x <- seq(from=0,to=50,length=100)` and `c<-50`. 
    Why are they called mirror image?


## Fitting

* Once we are given the knots and hinge functions, the fit can
be obtained by least squares.

\footnotesize

```{r}
Wage <- mutate(Wage,h1=pmax(0,38-age),h2=pmax(0,age-63))
ff <- lm(wage ~ h1+h2,data=Wage)
summary(ff)$coefficients
```

## Questions

* How do we choose the knots?
* What happens when there are multiple explanatory variables, and 
we allow for interactions between them?