---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: "MARS, part IV"
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
    fig_width: 4
    fig_height: 2.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE,message=FALSE)
```

## More details on the implementation

* Input
* Pre-processing
* Output
* Methods

## Software arguments/inputs

* Formula interface to specify response and explanatory variables.
* data argument for input data
* object of class `mars.control` that is a list comprised
of elements
`Mmax`, `d` and `trace`. `Mmax` is for the forward, `d` is for
the backward and `trace` is to print details of 
the fitting process.
    * Write a constructor, validator and helper
    function for this class. The helper should have
    defaults. If the user does not specify a 
    `mars.control` object on input, the helper
    should create one with the defaults.

## Pre-processing example: `lm()` vs `lm.fit()`

* `lm()` does a bit of pre-processing, incuding
setup of the design matrix `x` and response variable `y`,
and then calls `lm.fit()` to do the fitting.
* What we have so far in `fwd_selection()` and 
`bwd_selection()` amounts to an implementation 
of `lm.fit()`. 
* We need code to 
transform the formula, data, and control parameters
into the inputs to `fwd_selection()`.

## `lm()`

* `lm()` is very flexible and can read data from 
a data argument or the calling environment.

\tiny

```{r}
function (formula, data, subset, weights, na.action, method = "qr", 
    model = TRUE, x = FALSE, y = FALSE, qr = TRUE, singular.ok = TRUE, 
    contrasts = NULL, offset, ...) 
{
    ret.x <- x
    ret.y <- y
    cl <- match.call()
    mf <- match.call(expand.dots = FALSE)
    m <- match(c("formula", "data", "subset", "weights", "na.action", 
        "offset"), names(mf), 0L)
    mf <- mf[c(1L, m)]
    mf$drop.unused.levels <- TRUE
    mf[[1L]] <- quote(stats::model.frame)
    mf <- eval(mf, parent.frame())
    if (method == "model.frame") 
        return(mf)
    else if (method != "qr") 
        warning(gettextf("method = '%s' is not supported. Using 'qr'", 
            method), domain = NA)
    mt <- attr(mf, "terms")
    y <- model.response(mf, "numeric")
    w <- as.vector(model.weights(mf))
    if (!is.null(w) && !is.numeric(w)) 
        stop("'weights' must be a numeric vector")
    offset <- model.offset(mf)
    mlm <- is.matrix(y)
    ny <- if (mlm) 
        nrow(y)
    else length(y)
    if (!is.null(offset)) {
        if (!mlm) 
            offset <- as.vector(offset)
        if (NROW(offset) != ny) 
            stop(gettextf("number of offsets is %d, should equal %d (number of observations)", 
                NROW(offset), ny), domain = NA)
    }
    if (is.empty.model(mt)) {
        x <- NULL
        z <- list(coefficients = if (mlm) matrix(NA_real_, 0, 
            ncol(y)) else numeric(), residuals = y, fitted.values = 0 * 
            y, weights = w, rank = 0L, df.residual = if (!is.null(w)) sum(w != 
            0) else ny)
        if (!is.null(offset)) {
            z$fitted.values <- offset
            z$residuals <- y - offset
        }
    }
    else {
        x <- model.matrix(mt, mf, contrasts)
        z <- if (is.null(w)) 
            lm.fit(x, y, offset = offset, singular.ok = singular.ok, 
                ...)
        else lm.wfit(x, y, w, offset = offset, singular.ok = singular.ok, 
            ...)
    }
    class(z) <- c(if (mlm) "mlm", "lm")
    z$na.action <- attr(mf, "na.action")
    z$offset <- offset
    z$contrasts <- attr(x, "contrasts")
    z$xlevels <- .getXlevels(mt, mf)
    z$call <- cl
    z$terms <- mt
    if (model) 
        z$model <- mf
    if (ret.x) 
        z$x <- x
    if (ret.y) 
        z$y <- y
    if (!qr) 
        z$qr <- NULL
    z
}

```

## Pre-processing: model frames

* We'll simplify ...
* The data frame and formula are bundled into a 
"model frame", which is the data frame plus a 
terms attribute.
* R has tools for extracting the response and design
matrix from a model frame.

\footnotesize

```{r}
mars <- function(formula,data,control=NULL...) {
  cc <- match.call() # save the call
  mf <- model.frame(formula,data)
  y <- model.response(mf)
  mt <- attr(mf, "terms")
  x <- model.matrix(mt, mf)
  # if(is.null(control)) call the helper with default values
  # Then you are ready to go ...
}
```


## Value/output

* object of S3 class `mars`.
* inherits from class `lm` and includes
all of the components of the `lm()` from the final fit
    * Use `c()` to combine these with any of your
    own components.
* include `splits` data structure from final fit.
* write a constructor for this class -- no need 
for a validator or helper since you are the only
one who will call the constructor.
    
## Methods

* Use `methods()` to find a list of methods implemented
for the S3 class `lm`.
* Write more informative `print` and `summary` methods
for `mars` objects
* Write a `plot` method. 
    * The details are up to you, but you 
    should consult Section 3.5 of the Friedman paper (ANOVA decomposition).
    * Two sources of inspiration are the `plot.earth` method
    for `earth` objects (see the earth package), and `plot.Gam` for
    plotting generalized additive model components 
    (see the gam package).
* Write a `predict` method with the same
interface as `predict.lm`.
* `residuals()`, `fitted()`, `hatvalues()` and others
that depend only on the final `lm` can be used as-is
