---
title: 'Statistics 360:  Advanced R for Data Science'
subtitle: 'Lecture 7'
author: "Brad McNeney"
output: 
  beamer_presentation:
    includes:
      in_header: header_pagenum.tex
    toc: true
---

## Introduction to object-oriented programming in R

* Reading: Text, Chapters 12 and 13
* Topics:
    * general comments on object-oriented programming (OOP)
    * terminology
    * base objects vs OO objects
    * OOP with "S3" in R

## Object-oriented vs functional programming

* OOP aims to break a problem down into components
that are represented by "objects".
    * An object contains its data and the functions, or "methods"
    that can act on that data
* R is predominantly a functional programming language:
we break a problem down into functions.
* The debate over which stlye of programming is the best rages on.
    * Google "functional versus object-oriented programming" for a sample
* Practical note: algorithms and data structures go hand-in-hand for solving 
complex problems, and formalizing your data structure as an object is useful.

## OOP in R 

* Much of R uses OOP in some form, so if you want to contribute to,
extend, or just understand someone else's code you should learn
a little OOP.
* We will discuss three OOP systems: S3, R6 and S4.
    * S3 is the simplest and most widely used. Our text calls it
    "functional" OOP. 
    * RC and R6 are traditional "encapsulated" OOP systems that
    looks less familiar to R users but more familiar to programmers
    * S4 is a more formal version of S3
* Our goal is to learn a little about each style so that we can 
understand code written by others.
* For our project we will use S3.
    * Our MARS function will output an S3 object, and we will write
    print, summary, plot and other methods to make the interface
    familiar to R users.

## Why use S3?

* Very minimalist and flexible.
* Widely used, so others can understand your code.
* OOP can evolve as you work. 
    * Just like R is good for prototyping functional algorithms,
    S3 is good for prototyping OOP methods.
    * If your code needs more structure (e.g., you are
    starting to open it up to collaborators), you
    can formalize your OOP then.
* It "looks like R"

## Terminology: polymporhism and encapsulation

* **polymorphism:** As we've seen, functions like plot are generic and behave differently
when given different inputs; this is called polymorphism.
* **encapsulation:** We hide the details of an object behind an
interface.
    * Encapsulated OOP formally bundles of data and methods used to
    set and get data values.
    * Functional OOP provides functions that the user *should* use
    to get and set data.
    * Encapsulation allows the developer to change the implementation of 
    the object without breaking other code: just change the 
    relevant methods for getting and setting data.

## Example

\small

```{r}
data(mtcars)
ff <- lm(mpg~disp,data=mtcars)
class(ff)
# names(ff)
# ff$residuals
# residuals(ff)
# summary(ff)
```


## Terminology, continued

* We have been using the terms **class** and **method**.
* Other important terms:
    * **fields:** are the data of the class
    * **inheritance:** Classes can be organized in a hierarchy that we 
    search for an appropriate method. If a method does not exist for
    a child, or sub-class and we use the method from the parent, or super-class,
    then the child is said to inherit behaviour from the parent. 
   * **method dispatch** is the process of finding an appropriate method
   for a given class


## Base objects vs OO objects

* We have been using "object" to describe
data and functions in R in general.
* Now distinguish between "base" objects, such as numeric vectors,
and OO objects that have a class attribute.

\small

```{r}
x <- 1:4
attr(x,"class") # compare with class(x) -- misleading
attr(mtcars,"class")
attr(ff,"class")
```

## Base types

* Recall that base objects have a type. 
* There are 25 base types. 
* These are more about the underlying implementation of the
base object in memory, rather than the basis of an OOP system
* See the text, section 12.3 for more details on base object
types.


## OOP with S3

* S3 is an informal OO system and the most commonly-used.
    * E.g, it is the only OO system used in base R and the R stats package.
* Without strict rules, you have a lot of freedom, but can also write
bad code.
* We will discuss conventions for creating useful classes and methods
* We will use the `sloop` package recommended by the text to 
query objects about their class and available methods.

\small

```{r}
# install.packages("sloop")
library(sloop)
```


## S3 classes

* An S3 class is a base type with a class attribute

\small

```{r}
f <- factor(c("cat","dog","mouse"))
typeof(f)
attributes(f) # see also class(f) and inherits(f,"factor")
otype(f) # from sloop
s3_class(f) # from sloop
```

## Creating your own class

* Use `class()` to set the class after the object
has been created, or use `structure()`:

\small

```{r}
new_node <- function(data,childl=NULL,childr=NULL){
  structure(list(data=data,childl=childl,childr=childr),
                  class="node")
}
nn <- new_node(data=NULL) # Note: data should be a region object
s3_class(nn)
```

## Removing the class attribute

* We can simply remove the class with `attributes(f)$class <- NULL`
but it is better to use `unclass()`.

\small

```{r}
print(unclass(f))
otype(unclass(f))
```


## Class conventions

* No rules, but the text suggests a few conventions.

* Naming: Any string is OK, but stay away from `.`, which is 
the separator between generic and class names in naming methods.

* Constructor: Make a function named
`new_myclass()` to create an object with the correct 
structure.
* Validator: Make a function named `validate_myclass()` that
checks that the object's data makes sense, stops if not,
and otherwise returns the object.
* Helper: Make a function named `myclass()` that users
can use to create instances of the class. Should call
the validator, if one exists.

* Exercise (see week 7 exercises): 
Create validator and helper functions for the
node class example on the previous slide.

## Constructors

* The constructor should
    * Be called new_myclass().
    * Have one argument for the base object, and one for each attribute.
    * Check the type of the base object and the types of each attribute.

* I often write constructors whose base object is
a list, and my constructor has separate arguments for
each list element. 
    * See the `new_nod()` function, for example
    * This goes against the second of the 
above conventions. 
    
## Validator

* Write a validator if checking the validity of
the object's data may be computationally expensive.
* For your own code, where you are sure that you have
the arguments right, call the constructor. 
* Have users call the helper, which calls the validator.

## Helper

* As above, the helper can call the validator to 
check validity of data.
* Should have as many defaults as practical to 
make it easy to use.



## S3 generic functions and methods

* A generic function, like print, defines an 
interface (arguments) and finds an appropriate method
    * the method is an implementation specific 
    to the object class
    * finding an appropriate method is method dispatch

```{r}
print
ftype(print)
ftype(print.factor)
```


## Method dispatch: `UseMethod()`

* In simple cases, `UseMethod()` looks for  `generic.class()`,
and falls back on `generic.default()`. 
    * If neither exist, it throws an error.

* When a class inherits from a parent
class, the search gets more complicted.

## Example: print methods

* When you type the name of an object in the R console
you invoke print. 
* Different print methods exist for different classes of objects.
    * How many?

\small

```{r}
s3_methods_generic("print") # 208!
```

## Example: print.factor

* Most S3 methods are not exported from the packages in 
which they are defined, but you can view them
with sloop.

\tiny

```{r}
s3_get_method("print.factor")
```


## Writing methods when there is a generic

* Just write a function with name generic.class
    * See our `print.region()` method from lab 2.
* The method should have the same arguments
as the generic.
    * In the case of `print()` there is just one required argument, the object to be printed.

## Writing a generic

* Just need a call to `UseMethod()`

\small

```{r}
plot_regions <- function(x,...)   UseMethod("plot_regions")
plot_regions.tree <- function(tree){
  # set up empty plot
  plot(tree$data$x[,1],tree$data$x[,2],xlab="X1",ylab="X2") 
  plot_regions.node(tree$childl)
  plot_regions.node(tree$childr)
}
# add plot_regions.node(), recpart() then test
```

## Using inheritance

* Our MARS objects will contain the 
output of the final call to `lm()`. 
* Can make lm the parent class of our MARS objects.
    * Toy example:
    
\small

```{r}
new_mars <- function(formula,data) {
  ff <- lm(formula,data)
  structure(c(ff,list(ID="Hi, I'm a MARS object")),
            class=c("mars",class(ff)))
}
mm <- new_mars(mpg~cyl,data=mtcars)
s3_dispatch(print(mm))
```

## Further reading

* If you are interested in reading more about 
S3 classes on your own, see chapter 13 of the
text.
* Topics we skipped or skimmed:
    * Object styles (section 13.5)
    * Inheritance: NextMethod() and subclassing (section 13.6)
    * Dispatch details (section 13.7)